---
title: 'Midterm 2: Inference for numerical data'
output: html_document
---

```{r setup, include = FALSE}
# This is code to make your final report look nice! 
# Please don't modify these options! 
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")

library(tidyverse)
library(mosaic)
library(infer)
```

## The data

Every two years, the Centers for Disease Control and Prevention conduct the
Youth Risk Behavior Surveillance System (YRBSS) survey, where it takes data from
high schoolers (9th through 12th grade), to analyze health patterns. You will
work with a selected group of variables from a random sample of observations
during one of the years the YRBSS was conducted.

Run the code below to load the `yrbss` data set into your workspace.

```{r load-data}
YRBSS <- read_csv(here::here("data", 
                           "yrbss.csv"
                           )
                )
```

## Exploring the Data

```{r data-preview}
## Code to preview the dataset goes here!

```

__1. Write the code to preview the dataset.__

__2. Using the data preview, identify the cases in this data set.__


__3. How many cases are included in this sample?__


## Exploratory Data Analysis

We will first start by analyzing the weight of the participants in kilograms. 
This is recorded in the variable `weight`.

### Visualizing the Distribution of Weight 

__4. Create a histogram visualizing the distribution of weights for this sample.__
*Extra Credit:* Select a binwidth that makes the histogram look as continuous as
possible!

```{r weight-histogram}
## Code to make a histogram of the weights goes here!


```

### Summarizing Weights

__5. Using the favstats() function, find summary statistics for the weight for this sample.__

```{r weight-favstats}
## Code to find summary statistics of the weights goes here!


```

__6.  How many observations are we missing weights from?__


### Comparing Weights Across Levels of Exercise

Next, we will consider the possible relationship between a high schooler's
weight and their physical activity. To do this we will first create a new variable called `physical_3plus`, which will be coded as either "yes" if the student is physically active for *at least* 3 days a week, and "no" if not.

Run the code below to create this variable in the dataset!

```{r create new var}
YRBSS <- YRBSS %>% 
  mutate(physical_3plus = if_else(physically_active_7d > 2, "yes", "no")
         )
```

### Visualizing Distributions of Weight between Activity Groups

Now, we can use the `physical_3plus` variable to visualize the distribution of
weights between the two activity groups. 

__7. Complete the code below to created faceted histograms of the distribution of weights between the two physically active groups.__
*Extra Credit:* Select a binwidth that makes the histograms look as continuous
as possible!

```{r faceted-histograms-weight}
ggplot(data = YRBSS, 
       mapping = aes(x = ____)) + 
  ____ + 
  facet_wrap(~ ____)
```

__8. Describe the relationship between the amount of exercise a high schooler gets and their weight. Is this the relationship you expected to see?__


### Summarizing the Differences in Mean Weight

__9. Use the favstats() function to find summary statistics for the weight for each group of the `physical_3plus` variable.__

```{r weight-favstats-groups}
favstats(____ ~ ____, data = YRBSS)
```

__10. What is the observed difference in mean weight between the "yes" and "no" groups?__



## Statistical Inference

__11.  Write the hypotheses for testing if the average weights for high schoolers who exercise at least three days a week different from those who don't.__

*H0:* 

*HA:* 

__12. Evaluate if you believe any of the *three* conditions required for using the t-distribution to obtain our p-value are violated.__




__13. Complete the code below to find the p-value for testing if the average weights for high schoolers who exercise at least three days a week are different from those who don't.__

```{r t-test}
t_test(YRBSS, 
       response = ____, 
       explanatory = ____, 
       order = c("yes", "no"), 
       alternative = ____)
```

__14. Based on the p-value you obtained, what conclusion would you reach for the hypotheses you stated in #11?__


