---
title: 'Week 5 Day 1 Activity: Revisiting IMDb Movie Reviews --- Comparing Two Groups'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      fig.align = 'center', 
                      out.width = "70%")

library(tidyverse)
library(mosaic)

movies <- read_csv(here::here("Week 5 - Difference in Means & Midterm 1",
                              "activity", 
                              "day 1",
                              "Movies2016.csv")
                   )
```

This week our focus is comparing the means of two groups. With linear regression
we were able to compare the predicted mean response across different values of 
a continuous explanatory variable. This week, however, we are moving from a 
*continuous* explanatory variable to a **categorical** explanatory variable! 

### Movies released in 2016

Recall the dataset we used a couple of weeks ago, visualizing the distribution
of IMDB movie ratings. The dataset included eight variables, listed below.  

| **Variable** 	| **Description** |
|----	|-------------	|
| `movie_title` | Title of movie |
| `budget_mil` | Amount of money (in US $ millions) budgeted for the production of the movie |
| `revenue_mil` | Amount of money (in US $ millions) the movie made after release|
| `duration` | Length of the movie (in minutes)|
| `content_rating` | Rating of the movie (`G`, `PG`, `PG-13`, `R`, `Not Rated`)|
| `imdb_score` | IMDb user rating score from 1 to 10 |
| `genres` | Categories the movie falls into (e.g., Action, Drama, etc.) |
| `facebook_likes` | Number of likes a movie receives on Facebook |


## Comparing Two Groups

Let's compare the budget for movies rated PG and movies rated R. 
Technically, the dataset has four different content ratings: Not Rated, PG,
PG-13, and R. To make the comparison of PG and R movies easier, I went ahead
and made a smaller dataset containing only these two ratings.

In case you are interested, this is the code I used!

```{r}
movies_pg <- filter(movies, 
                    content_rating %in% c("PG", "R")
                    )
```


Similar to summarizing the mean of all of the movies, we have two options to
compare these two groups: 

1. Use summary statistics
2. Use visualizations

\newpage

## Summary Statistics

Let's start with summary statistics. Our familiar friend `favstats()` can help 
us compare summary statistics across different groups. Before when we used 
`favstats()` we only had one variable, but now we have two! 

Now, we will use two variables as a "formula", which looks like 
`response ~ explanatory`. So, the budget of the film is the response and the 
content rating is the explanatory variable. So, our code looks like: 

```{r stat-comparison, eval = FALSE}
favstats(budget_mil ~ content_rating, 
         data = movies_pg)
```

Use the output from the `favstats()` function to answer the following questions:

```{r, echo = FALSE}
favstats(budget_mil ~ content_rating,
         data = movies_pg)
```

\vspace{0.8in}

1. Report the mean budget amount for the PG rating.  Use appropriate notation.

\vspace{0.3in}

2. Report the mean budget amount for the R rating.  Use appropriate notation.

\vspace{0.3in}

3. Calculate the difference in mean budget amount for movies in 2016 with a PG
rating minus those with a R rating.  Use appropriate notation with informative
subscripts.

\vspace{0.8in}

4. Which content rating has the largest standard deviation? 

\vspace{0.8in}

5. Which content rating has the largest sample size? What effect does sample 
size have on each group's standard deviation?

\vspace{0.8in}

6.  Which content rating has the largest IQR?

\vspace{0.8in}

7. Based on the summary statistics, roughly sketch what you believe the
distribution of budgets for each content rating would look like. 

\vspace{2in}

## Visualizing Differences

Let's refresh ourselves on the different ways to plot a numerical variable. 

8. What are the three types of plots used to plot a single quantitative
variable?

\vspace{0.8in}

9. For each type of plot, how would you include a categorical variable in 
the plot? 

\pagebreak

## Faceted Histograms

When we want to add a categorical variable (like `content_rating`) to a
histogram, we create separate plots for each level of the categorical variable.
These separate plots are called **facets**. We are comparing the budgets for
PG and R movies, so we will have two facets, one per content rating. 

The code to make a faceted histogram looks like the following:

```{r facet-preview, eval = FALSE}
ggplot(data = movies_pg, 
       mapping = aes(x = budget_mil)) + 
  geom_histogram(binwidth = 100) + 
  labs(x = "Movie Budget (in millions)") +
  facet_wrap(~ content_rating)
```

Notice the last line is the only new part! That line creates a faceted plot 
(using `facet_wrap()`) and says to facet "by" (`~`) the `content_rating`. 

Let's look at what this plot ends up looking like:

```{r facet-hist, echo = FALSE}
ggplot(data = movies_pg, 
       mapping = aes(x = budget_mil)) + 
  geom_histogram(binwidth = 100) + 
  labs(x = "Movie Budget (in millions)") +
  facet_wrap(~ content_rating) 
```

10. How would you describe the shape of the distribution of PG movie budgets? 

\vspace{1in}

11. How would you describe the shape of the distribution of R movie budgets? 

\vspace{0.8in}

12. Visually, does it appear that there are outliers in either of these groups?

\vspace{0.8in}

13. How do these distributions compare to what you expected in your sketches
(in #7)?

\vspace{0.8in}

## Side-by-Side Boxplots

Another way we can incorporate a categorical into our plots is to plot our 
boxplots for each group side-by-side. As opposed to faceting, these boxplots
will be on the **same** plot. We only need to add one extra piece to our 
previous code: a categorical variable. 

Before, we either plotted our **one** numerical variable horizontally (using 
`x`) or vertically (using `y`). 

```{r one-box, echo = FALSE}
horizontal <- ggplot(data = movies, 
             mapping = aes(x = imdb_score)) + 
  geom_boxplot() + 
  labs(x = "IMDB Score")

vertical <- ggplot(data = movies, 
                   mapping = aes(y = imdb_score)) + 
  geom_boxplot() + 
  labs(y = "IMDB Score")

gridExtra::grid.arrange(horizontal, vertical, nrow = 1)
```

Now. we need to plot **two** boxplots side-by-side. Similar to before, we can
stack the plots horizontally or vertically. 

\pagebreak

### Horizontal Stacking

```{r horizontal-box}
ggplot(data = movies_pg, 
       mapping = aes(y = budget_mil, x = content_rating)) +
  geom_boxplot() +
  labs(x = "Content Rating of Movie", 
       y = "Movie Budget (in millions)")
```

14. Why are the boxplots stacked side-by-side horizontally? What part of the 
`R` code does this?

\pagebreak

### Vertical Stacking

```{r vertical-box}
ggplot(data = movies_pg, 
       mapping = aes(x = budget_mil, y = content_rating)) +
  geom_boxplot() +
  labs(y = "Content Rating of Movie", 
       x = "Movie Budget (in millions)")
```

15. How was the previous code changed to stack the boxplots side-by-side
vertically? 

\vspace{0.8in}
 
16. Which orientation do you prefer?
