---
title: "Final Exam Question Bank"
author: "Stat 218"
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - multicol
    - setspace
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(janitor)
library(pander)
library(infer)
library(praise)

set.seed(1234)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted. It is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours prior to the exam.

\vspace{0.5cm}

## Golden Ticket

| Scenario | One Categorical Response | Two Categorical Variables | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Bar plot | Side-by-Side Bar plot, Stacked Bar plot, Mosaic plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
| Summary measure | Proportion | Difference in Proportions | Mean | Slope or Correlation | Difference in Means |
| Parameter notation | $\pi$ | $\pi_1 - \pi_2$ | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
| Statistic notation | $\widehat{p}$ | $\widehat{p}_1 - \widehat{p}_2$ | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}


## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\mu_1 - \mu_2) = \sqrt{\frac{\sigma_1}{n_1} + \frac{\sigma_2}{n_2}}$$

\newpage 


## Final Exam Question Bank

### Wild Mushrooms

\noindent \textbf{Q1}[17 points]
Wild mushrooms, such as chanterelles or morels, are delicious, but eating wild
mushrooms carries the risk of accidental poisoning. Even a single bite of the
wrong mushroom can be enough to cause fatal poisoning. An amateur mushroom
hunter is interested in finding an easy rule to differentiate poisonous and
edible mushrooms. They think that the mushroom’s gills (the part which holds and
releases spores) might be related to a mushroom’s edibility. They used a data
set of 8124 mushrooms and their descriptions. For each mushroom, the data set
includes whether it is edible or poisonous and the spacing of the gills (Broad
or Narrow). 

**Please Note**: According to The Audubon Society Field Guide to North American
Mushrooms, there is no simple rule for determining the edibility of a mushroom;
no rule like "leaflets three, leave them be" for Poisonous Oak and Ivy.

```{r mushroom-data, include = FALSE}
mushrooms <- read_csv(here::here(
  "Week 10 - Review & Final", 
  "final", 
  "individual", 
  "data", 
  "mushrooms.csv")
  ) %>% 
  mutate(class = if_else(class == "e", "Edible", "Poisonous"), 
         `gill-size` = if_else(`gill-size` == "n", "Narrow", "Broad")
         )

```

```{r mushroom-table}
mushrooms %>% 
  group_by(class, `gill-size`) %>% 
  count() %>% 
  pivot_wider(names_from = `gill-size`, 
              values_from = n) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  pander()
```


\vspace{0.2cm}

\noindent
\textbf{(a)}[4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.2cm}

Whether the mushroom is edible or poisonous is the (explanatory / response) \underline{\hspace{3in}} and it is
(categorical / quantitative) \underline{\hspace{3in}}. 

\vspace{0.2cm}

Gill size (Broad or Narrow) is the (explanatory / response) 
\underline{\hspace{3in}} and it is (categorical / quantitative)
\underline{\hspace{3in}}. 

\vspace{0.2cm} 

\noindent
\textbf{(b)}[3 points] Calculate the proportion of mushrooms with a broad gill
size that are poisonous. *Give appropriate notation with informative subscripts
if necessary.* 

\vspace{0.2cm}

**Work:**

\vspace{1cm}

**Answer:** \underline{\hspace{3in}}

\vspace{0.2cm}

**Notation:** \underline{\hspace{3in}}

\vspace{0.2cm}

\noindent
\textbf{(c)}[3 points] Calculate the proportion of mushrooms with a narrow gill
size that are poisonous. *Give appropriate notation with informative subscripts
if necessary.* 

\vspace{0.2cm}

**Work:**

\vspace{1cm}

**Answer:** \underline{\hspace{3in}}

\vspace{0.2cm}

**Notation:** \underline{\hspace{3in}}

\vspace{0.2cm}

\noindent
\textbf{(d)}[2 points] Using your answers to (b) and (c), fill in the correct
names next to each color, to label the bar chart showing the relationship
between gill size (broad or narrow) and whether the mushroom is edible. 

```{r mushroom-barplot}
ggplot(data = mushrooms, 
       mapping = aes(x = `gill-size`, fill = class)
       ) +
  geom_bar() + 
  labs(x = "Size of Mushroom Gills", 
       y = "Number of Mushrooms", 
       fill = "Class of Mushroom") +
  theme(legend.text = element_blank())

```

\vspace{0.25in}

\noindent
\textbf{(d)}[3 points] Based on the plot, describe the relationship between a 
mushrooms gill size and whether it is edible or not. 

\vspace{1in}

<!-- Add question about shuffling under H0 -->

\noindent
\textbf{(e)}[2 points] Suppose the Chi-Squared test resulted in a "significant"
p-value. Which of the following would be the correct scope of inference for this
study?

\begin{itemize}
\item[(i)] It can be inferred for all mushrooms that gill size causes a mushroom
to be poisonous. 
\item[(ii)] It can be inferred for all mushrooms that gill size is associated
with whether a mushroom is poisonous. 
\item[(iii)] It can be inferred for this sample of mushrooms that gill size
causes a mushroom to be poisonous. 
\item[(iv)] It can be inferred for this sample of mushrooms that gill size is
associated with whether a mushroom is poisonous.
\end{itemize}

\newpage

`r praise()` Congratulations on completing STAT 218! 

Have a great summer!




### Mendelay Genetics

### Professor Attractiveness

Data were scraped from _ratemyprofessors.com_, a website which allows college
and university students to assign ratings to professors and campuses of
American, Canadian, and United Kingdom institutions. In order for the rating to
be posted, a rater must rate the professor in the following categories: 
`overall quality`, `level of difficulty`, and their `hotness`.  

The Dean of the College of Science and Math would like to know if the difficulty
rating of female professor's at Cal Poly is related to how attractive they are
perceived to be. 

A table of observed counts in shown below.  

```{r, professor-table, echo = FALSE}
professorss <- tibble(attractiveness = 
                       c(rep("Attractive", 172), 
                         rep("Unattractive", 314)
                         ), 
                     difficulty = 
                       c(rep("Easy", 76), 
                         rep("Medium", 51), 
                         rep("Hard", 45), 
                         rep("Easy", 54), 
                         rep("Medium", 90), 
                         rep("Hard", 170)
                         )
                     )
professors %>% 
  group_by(attractiveness, difficulty) %>% 
  count() %>% 
  rename(Difficulty = difficulty) %>% 
  pivot_wider(names_from = attractiveness, 
              values_from = n) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  pander()

```

\noindent \textbf{(a)}[3 points] What type of test should be performed with
these data to answer the research question?  

\vspace{1in}

\noindent \textbf{(b)} [3 points] When analyzing the data, the administrators 
chose to use a $\chi^2$ distribution to obtain their p-value. Was this an
appropriate choice? Why or why not? 

\vspace{1in}

\noindent \textbf{(c)} [4 points] Write out the null and alternative hypotheses
for the test you stated in __(a)__. 

\noindent $H_0:$ 

\vspace{0.5in}

\noindent $H_A:$ 

\vspace{0.5in}

\noindent \textbf{(d)} [2 points] What distribution should be used to find the
p-value for the hypothesis test above?

\vspace{0.25in}

\noindent \textbf{(e)} [4 points] A p-value less than 0.000000142 ($X^2$
statistic = 50) was found using the distribution stated in __(d)__. Based on 
the p-value obtained, report your conclusion to the hypothesis test
__in the context of the problem.__ 

\vspace{1in}


### Guilty Decisions

\noindent
\textbf{Q3}[11 points] 
The effect of guilt on how a decision maker focuses on a problem was
investigated in the *Journal of Behavioral Decision Making* (January 2007). A
total of 155 volunteer students participated in the study, where each was
randomly assigned to one of three emotional states (guilt, anger, or neutral)
through a reading / writing task. Immediately after the task, the students were
presented with a decision problem (e.g., whether or not to spend money on
repairing a very old car). The researchers found that a higher proportion of
students in the guilty-state group chose not to repair the car than those in the neutral-state and anger-state groups.

\noindent
\textbf{(a)}[3 pts] What is the study design? Explain your reasoning. Select
one.

\begin{itemize}
\item[(i)] Observational study. The researchers did not take a random sample of
students.
\item[(ii)] Observational study. There is no random assignment of students to 
emotional state.
\item[(iii)] Experiment. The students are a representative sample of all
students.
\item[(iv)] Experiment. The students were randomly assigned to emotional state.
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{(b)}[3 pts] Which types of sampling bias may be present in
this study? Select all that are present, or if you believe there is no bias 
present, select option 4) No bias.

\begin{itemize}
\item[(i)] Selection bias
\item[(ii)] Non-response bias
\item[(iii)] Response bias
\item[(iv)] No bias
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{(c)}[3 pts] The researchers found a difference in the 
proportion of students in the guilty-state group chose not to repair the car
than those in the neutral-state and anger-state groups. Can we conclude that the emotional state group caused a higher proportion of students to not repair the
car? Select one.

\begin{itemize}
\item[(i)] Yes, because these data are from a representative sample.
\item[(ii)] No, because the students are volunteers.
\item[(iii)] Yes, because the researchers evened out confounding variables
across emotional state group.
\item[(iv)] No, because the sample size is not large enough.
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{(d)} [2 pts] Which type of plot would be the most appropriate
to display the relationship between each level of emotional state and whether
students choose to repair the car? Select all that apply.

\begin{itemize}
\item[(i)] Scatterplot
\item[(ii)] Filled bar plot
\item[(iii)] Pie chart
\item[(iv)] Side-by-side boxplot
\end{itemize}


## Distribution of Fish on Blackfoot River

```{r fish-data}

fish <- read_csv(here::here(
  "Week 10 - Review & Final", 
  "final", 
  "individual", 
  "data", 
  "BlackfootFish.csv")
  ) %>% 
  drop_na(section, species) %>% 
  mutate(species = case_when(species == "Brown" ~ "Brown Trout", 
                             species == "Bull" ~ "Bull Trout", 
                             species == "RBT" ~ "Rainbow Trout", 
                             species == "WCT" ~ "Whitefish"), 
         year = as.factor(year)
         )

```

We have data on fish caught in the Blackfoot River near Helena, Montana by
Montana Fish, Wildlife, & Parks personnel over a number of years. They used electrofishing equipment to attract the fish to the boat, then dipped them out
of the water with nets, measured length in cm and weight in grams. They are
often working in cold conditions in late autumn or early spring, so some
measurement error is expected. 

These data are not from a random sample. The goal is to catch all fish within a
section of the Blackfoot River every few years to assess the health of the
population. Montana Fish, Wildlife, & Parks is interested in analyzing if the
prevalence of the different species of fish has stayed relative stable over 
time.  

```{r fish-preview}
fish %>% 
  select(length, weight, year, section, species) %>% 
  sample_n(10)
```

\noindent \textbf{(a)} [2 points] Based on the output above, what is the
observational unit for this study?

\vspace{0.5in}

\noindent \textbf{(b)} [3 points] Based on the output above, what type of 
variable is `year`? Given the stated analysis, is this the correct data type for
this variable?

\vspace{1in}

\noindent \textbf{(c)}

\noindent \textbf{(d)}

```{r fish-table}
fish %>% 
  group_by(year, species) %>% 
  count() %>% 
  pivot_wider(names_from = year, 
              values_from = n) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  pander()
```

```{r fish-plot}
fish %>% 
  ggplot(
    mapping = aes(y = year, fill = species)
       ) +
  geom_bar(position = "fill") +
  labs(y = "Species of Fish on Blackfoot River", 
       x = "Proportion of Caught Fish") +
  theme(legend.position = "top")
```


\noindent 


## General Concepts

\noindent \textbf{}[2 points] Suppose you reject the null hypothesis at the 0.05
level of significance. A colleague had planned to use a 0.01 level of
significance instead. Will your colleague also reject the null hypothesis?

\begin{itemize}
\item[(a)] Yes
\item[(b)] No
\item[(c)] Maybe
\item[(d)] Changing a level of significance cannot affect decisions.
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{}[2 points] In hypothesis testing,

\begin{itemize}
\item[(a)] a type II error occurs when you fail to reject a false null
hypothesis.
\item[(b)] you do not need to decide on a level of significance before you find
a p-value; you can adjust based on the p-value you observe.
\item[(c)] one of the main sources of type I errors is that the sample size was
not large enough.
\item[(d)] p-values are the probability that the null hypothesis is false.
\end{itemize}
