---
title: "Lab 4: COVID Immunization and Infection Rates"
author: "Your name here!"
output: html_document
---

As a reminder, answer **every** question included in the lab. The questions that
are boldface are required to be answered in Canvas. 

## Getting set-up

We will use three packages this week, two we have used before (**tidyverse**
and **infer**) and one new package (**moderndive**). The moderndive package 
has tools to help us obtain the coefficients, standard errors, and t-statistics
from our regression line. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE)
library(tidyverse)
library(infer)
library(moderndive)
```


### COVID immunization and infection rates

According to the *Washington Post* "States with higher vaccination rates now 
have markedly fewer coronavirus cases, as infections are dropping in places
where most residents have been immunized and are rising in many places people
have not." [@keating2021] In this article they found that there are differences
in infection rates for different counties within a specific state. To check this
claim, a random sample of 125 counties from different states was assessed. 
Vaccination rates and number of cases per 100,000 residents were found for each
county.  Researchers want to assess if counties with a high vaccination rate
tend to have lower coronavirus case rates.

## Loading in data

As usual, we'll load in the data we are going to work with: `GSS_clean.csv`.
The data should be inside the `data` folder in your RStudio Cloud. Just like
before, we'll use the `read_csv()` function to read in the data. 

```{r data-load}
covid <- read_csv(here::here("Week 4 - SLR", 
                           "lab",
                           "data", 
                           "covid_vaccinations.csv"
                           )
                )


covid <- read_csv(here::here("data", 
                           "covid_vaccinations.csv"
                           )
                )
```


#### Summarize and visualize the data {-} 

To find the correlation between the variables, `PercentImmunized` and `Case_per_100K` highlight and run lines 10--13 in the `R` script file.

```{r, echo=TRUE, eval=FALSE}
covid %>% 
  select(c("PercentImmunized", "Case_per_100K")) %>%
  cor(use="pairwise.complete.obs") %>%
  round(3)
```

__1.  Report the value of correlation between the variables.__ 

__2. Calculate the value of the coefficient of determination between `PercentImmunized` and `Case_per_100K`.__

3. Interpret the value of the coefficient of determination in context of the
problem.

In the next part of the activity we will assess the linear regression between
percent immunized and cases per 100,000.  

Enter the variable `Case_per_100K` for `response` and the variable `PercentImmunized` for `explanatory` in line 17.  

```{r line}
# Fit linear model: y ~ x
covid_lm <- lm(response~explanatory, data=covid)

get_regression_table(covid_lm)
```

4. Give the value of the slope of the regression line.  Interpret this value in context of the problem.
\vspace{0.6in}

#### Conditions for the least squares line {-}

Highlight and run lines 21--34 to produce the diagnostic plots needed to assess conditions to use theory-based methods.  Use the scatterplot and the residual plots to assess the validity conditions for approximating the data with the $t$-distribution.

```{r, echo=TRUE, eval=FALSE, warning=FALSE}
#Scatterplot
covid %>% # Pipe data set into...
  ggplot(aes(x = PercentImmunized, y = Case_per_100K))+  # Specify variables
  geom_point() +  # Add scatterplot of points
  labs(x = "Percent Immunized",  # Label x-axis
       y = "Number of cases per 100K",  # Label y-axis
       # Be sure to tile your plots
       title = "Scatterplot of Percent Immunized vs. Infection Rate of COVID in US Counties") + 
  geom_smooth(method = "lm", se = FALSE)  # Add regression line

#Diagnostic plots
covidLM <- lm(Case_per_100K~PercentImmunized, data = covid) # Fit linear regression model
par(mfrow=c(1,2)) # Set graphics parameters to plot 2 plots in 1 row
plot(covidLM, which=1) # Residual vs fitted values
hist(covidLM$resid, xlab="Residuals", ylab="Frequency",
     main = "Histogram of Residuals") # Histogram of residuals
```

5. **Are the conditions met to use the $t$-distribution to approximate the sampling distribution of the standardized statistic? Justify your answer.**

\vspace{1.5in}

#### Ask a research question {-}

6. Write out the null and alternative hypotheses in notation to test *correlation* between the percent immunized in US counties and the infection rate.

|    $H_0:$

|    $H_a:$

#### Use statistical inferential methods to draw inferences from the data {-}

##### Hypothesis test {-}

Use the `regression_test()` function in `R` (in the `catstats` package) to simulate the null distribution of sample **correlations** and compute a p-value.  We will need to enter the response variable name and the explanatory variable name for the formula, the data set name (identified above as `covid`), the summary measure used for the test, number of repetitions, the sample statistic (value of correlation), and the direction of the alternative hypothesis.

The response variable name is `Case_per_100K` and the explanatory variable name is `PercentImmunized`.

7. What inputs should be entered for each of the following to create the simulation to test correlation?

\vspace{.5 mm}

* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.2in}

* Summary measure (choose `"slope"` or `"correlation"`):

\vspace{.2in}
* As extreme as (enter the value for the sample slope):

\vspace{0.2in}

* Number of repetitions:
    
\vspace{.2in}

Using the `R` script file for this activity, enter your answers for question 7 in place of the `xx`'s to produce the null distribution with 1000 simulations.  Highlight and run lines 37--42.  **Upload a copy of your plot showing the p-value to Gradescope for your group.**

```{r, echo=TRUE, eval=FALSE}
regression_test(Case_per_100K~PercentImmunized, # response ~ explanatory
               data = covid, # Name of data set
               direction = "xx", # Sign in alternative ("greater", "less", "two-sided")
               summary_measure  = "xx", # "slope" or "correlation"
               as_extreme_as = xx, # Observed slope or correlation
               number_repetitions = 1000) # Number of simulated samples for null distribution
       
```

8.  Report the p-value from the `R` output. 
\vspace{0.3in}

9. Interpret the p-value in context of the problem.
\vspace{0.8in}

#### Simulation-based confidence interval {-}

We will use the `regression_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of sample **correlations** and calculate a confidence interval. Fill in the `xx`'s in the the provided `R` script file to find a 90\% confidence interval. Highlight and run lines 45--49. 

```{r, echo=TRUE, eval=FALSE}
regression_bootstrap_CI(Case_per_100K~PercentImmunized, # response ~ explanatory
   data = covid, # Name of data set
   confidence_level = xx, # Confidence level as decimal
   summary_measure = "xx", # Slope or correlation
   number_repetitions = 1000) # Number of simulated samples for bootstrap distribution
```
10.  Report the bootstrap 90\% confidence interval in interval notation.  
\vspace{0.5in}

11. Interpret the 90\% confidence interval in context of the problem.
\vspace{0.8in}
   
#### Communicate the results and answer the research question {-}

12. Based on the p-value, write a conclusion in context of the problem.

\vspace{.8in}

13. Using a significance level of 0.1, what decision would you make?
\vspace{0.2in}

14. What type of error is possible?
\vspace{0.3in}

15. Interpret this error in context of the problem.
\vspace{0.8in}

16. Write a paragraph summarizing the results of the study as if you are reporting these results in your local newspaper.  **Upload a copy of your paragraph to Gradescope for your group.**  Be sure to describe:

* Summary statistic

* Test statistic and interpretation

* P-value and interpretation

* Confidence interval and interpretation

* Conclusion (written to answer the research question)

* Scope of inference

\vspace{3in}

\newpage