---
title: 'Week 4 Day 2: Golf Driving Distance'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(infer)

set.seed(1234)

my_theme <- theme(axis.text = element_text(size = 16), 
                  axis.title = element_text(size = 18)
                  )

```


### Terminology review

In this week's in-class activity, we will use theory-based methods for hypothesis tests and confidence intervals for a linear regression slope. Some terms covered in this activity are:

* Slope 

* Regression line

To review these concepts, see Chapters 3 and 7 in the textbook.


### Golf driving distance

In golf the goal is to complete a hole with as few strokes as possible.  A long driving distance to start a hole can help minimize the strokes necessary to complete the hole, as long as that drive stays on the fairway.  Data was collecting on 354 PGA and LGPA players in 2008 [@pga].  For each player, the average driving distance (yards), fairway accuracy (percentage), and sex was measured.  Use these data to assess, "Does a professional golfer give up accuracy when they hit the ball farther?"

```{r, echo = TRUE, collapse=FALSE}
# Read in data set
golf <- read.csv("https://math.montana.edu/courses/s216/data/golf.csv")
```

#### Plot review. {-}

Use the provided `R` script file to create a scatterplot to examine the relationship between the driving distance and percent accuracy by filling in the variable names (`Driving_Distance` and `Percent_Accuracy`) for `xx` and `yy` in line 10.  Highlight and run lines 1--16. 
 
```{r, echo=TRUE, eval=FALSE}
golf %>% # Pipe data set into...
ggplot(aes(x = xx, y = yy))+  # Specify variables
  geom_point() +  # Add scatterplot of points
  labs(x = "Driving Distance",  # Label x-axis
       y = "Percent Accuracy",  # Label y-axis
       title = "Scatterplot of Driving Distance by Percent Accuracy") + 
               # Be sure to tile your plots
  geom_smooth(method = "lm", se = FALSE)  # Add regression line
```

1. Sketch the plot created below. Based on your plot, does it appear that there is a relationship between driving distance and percent accuracy? Note: `Driving Distance` should be on the $x$-axis.

\vspace{2in}

#### Conditions for the least squares line {-}

When performing inference on a least squares line, the follow conditions are generally required:

* *Independent observations* (for both simulation-based and theory-based methods): individual data points must be independent.
    - Check this assumption by investigating the sampling method and determining if the observational units are related in any way.
    
* *Linearity* (for both simulation-based and theory-based methods): the data should follow a linear trend.
    - Check this assumption by examining the scatterplot of the two variables, and a scatterplot of the residuals (on the $y$-axis) versus the fitted values (on the $x$-axis). The pattern in the residual plot should display a horizontal line.

* *Constant variability* (for theory-based methods only): the variability of points around the least squares line remains roughly constant
    - Check this assumption by examining a scatterplot of the residuals (on the $y$-axis) versus the fitted values (on the $x$-axis). The variability in the residuals around zero should be approximately the same for all fitted values.

* *Nearly normal residuals* (for theory-based methods only: residuals must be nearly normal.
    - Check this assumption by examining a histogram of the residuals, which should appear approximately normal^[A better plot for checking the normality assumption is called a *normal quantile-quantile plot* (or QQ-plot). However, this type of plot will be covered in a future course].

\newpage 

The scatterplot generated in question 1 and the residual plots shown below will be used to assess these conditions for approximating the data with the $t$-distribution.

```{r}
lm.golf <- lm(Percent_Accuracy~Driving_Distance, data = golf) # Fit linear regression model
par(mfrow=c(1,2)) # Set graphics parameters to plot 2 plots in 1 row
plot(lm.golf, which=1) # Residual vs fitted values
hist(lm.golf$resid, xlab="Residuals", ylab="Frequency",
     main = "Histogram of Residuals") # Histogram of residuals
```
2. Are the conditions met to use the $t$-distribution to approximate the sampling distribution of the standardized statistic? Justify your answer.

\vspace{1.5in}


#### Ask a research question {-}

3. Write out the null hypothesis in words.

\vspace{1in}

4. Using the research question, write the alternative hypothesis in notation to test the slope.

\newpage

#### Summarize and visualize the data {-}

Using the provided `R` script file, enter the response variable name, `Percent_Accuracy`, into the `lm()` (linear model) function for `response` and the explanatory variable name, `Driving_Distance`, for `explanatory` in line 26 to get the linear model output.  Highlight and run lines 26--27.

```{r, echo=TRUE, eval =FALSE}
lm.golf <- lm(response~explanatory, data=golf) # lm(response~explanatory)
round(summary(lm.golf)$coefficients, 5)
```

5.  Using the output from the evaluated `R` code above, write the equation of the regression line in the context of the problem using appropriate statistical notation.

\vspace{1in}

6.  Interpret the estimated slope in context of the problem.

\vspace{1in}

#### Use statistical inferential methods to draw inferences from the data {-}

##### Hypothesis test {-}
To find the value of the standardized statistic to test the slope we will use, 

$$
T = \frac{\mbox{slope estimate}}{SE} = \frac{b_1}{SE(b_1)}.
$$
 
We will use the linear model `R` output above to get the estimate for slope and the standard error of the slope.

7.  What are the values of $b_1$ and $SE(b_1)$?  Where in the linear model `R` output can you find these values?
\vspace{0.5in}

8.  Calculate the standardized statistic for slope.  Identify where this calculated value is in the linear model `R` output.

\newpage

9.  Interpret the standardized statistic in context of the problem.

\vspace{0.8in}

10.  The p-value in linear model `R` output is the two-sided p-value for the test of significance for slope.  Report the p-value to answer the research question.  

\vspace{0.5in}

11. Based on the p-value, how much evidence is there against the null hypothesis?

\vspace{0.5in}

##### Confidence interval {-}
Recall that a confidence interval is calculated by adding and subtracting the margin of error to the point estimate.  
$$\mbox{point estimate}\pm t^*SE(\mbox{estimate}).$$
When the point estimate is a regression slope, this formula becomes 
$$b_1 \pm t^* SE(b_1).$$
 
The $t^*$ multiplier comes from a $t$-distribution with $n-2$ degrees of freedom.  Recall for a 95\% confidence interval, we use the 97.5\% percentile (95\% of the distribution is in the middle, leaving 2.5\% in each tail).  The sample size for this study is 354 so we will use the degrees of freedom 352 ($n-2$).

```{r, echo=TRUE, collapse=FALSE}
qt(0.975, 352) # 95% t* multiplier 
```

12. Calculate the 95\% confidence interval for the true slope.
\vspace{0.8in}

13. Interpret the 95\% confidence interval in context of the problem.

\vspace{.8in}

#### Communicate the results and answer the research question {-}

14. Write a conclusion to answer the research question in context of the problem.

\vspace{.8in}

### Multivariate plots {-}

Another variable that may affect the percent accuracy is the sex of the golfer. We will look at how this variable may change the relationship between driving distance and percent accuracy. Highlight and run lines 33--39 to produce the multivariate plot. 

```{r, out.width="60%", echo=TRUE, eval=FALSE}
golf %>%
  ggplot(aes(x = Driving_Distance, y = Percent_Accuracy, color=Gender))+  # Specify variables
  geom_point(aes(shape = Gender), size = 3) +  # Add scatterplot of points
  labs(x = "Driving Distance (m)",  # Label x-axis
       y = "Percent Accuracy",  # Label y-axis
       color = "Sex", shape = "Gender",
       # Be sure to title your plots
       title = "Scatterplot of Golf Driving Distance and Percent Accuracy by Sex") + 
  geom_smooth(method = "lm", se = FALSE)  # Add regression line
```

15. Does the association between driving distance and percent accuracy change dependent on sex of the golfer?  Explain your answer.
\vspace{1in}

16.  Explain the impact sex has on the two variables.
\newpage

### Take-home messages


1. To check the validity conditions for using theory-based methods we must use the residual diagnostic plots to check for normality of residuals and constant variability, and the scatterplot to check for linearity.  

2. To interpret a confidence interval for the slope, think about how to interpret the sample slope and use that information in the confidence interval interpretation for slope.  

3. Use the explanatory variable row in the linear model `R` output to obtain the slope estimate (`estimate` column) and standard error of the slope (`Std. Error` column) to calculate the standardized slope, or T-score.  The calculated T-score should match the `t value` column in the explanatory variable row. The standardized slope tells the number of standard errors the observed slope is above or below 0.

4. The explanatory variable row in the linear model `R` output provides a **two-sided** p-value under the `Pr(>|t|)` column.

5. The standardized slope is compared to a $t$-distribution with $n-2$ degrees of freedom in order to obtain a p-value.  The $t$-distribution with $n-2$ degrees of freedom is also used to find the appropriate multiplier for a given confidence level.


### Additional notes

Use this space to summarize your thoughts and take additional notes on this week's activity and material covered.

\newpage