---
title: "Visualizing Categorical Variables"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(message = FALSE, 
                      fig.align = "center", 
                      echo = TRUE)

## Load in the packages necessary for our analysis: tidyverse


```

## Nightlight use and myopia 

In a study reported in *Nature* (Quinn, 1999), a survey of 479 children found
that those who had slept with a nightlight or in a fully lit room before the age
of two had a higher incidence of nearsightedness (myopia) later in childhood.

In this study, there were two variables studied:

- `Light`: level of light in room at night (no light, nightlight, full light)
- `Sight`: level of myopia developed later in childhood (high myopia, myopia,
no myopia).  

__1. Which variable is the explanatory variable? Which is the response variable?__


Let's load the data in! 

```{r data, include = FALSE}
myopia <- read_csv(here::here("Week 7 - ANOVA & One Proportion", 
                              "lab", 
                              "data", 
                              "ChildrenLightSight.csv")
                   ) 
```

__2. Click on the dataset to open it up in the data previewer. What is the observational unit for this study?__



## Visualizing One Categorical Variable

### Frequency Bar Plots 

If we wanted to know how many children in our data set were in each level of
myopia, we would create a frequency bar plot of the variable `Sight`. Creating
a bar plot is similar to what we've done before. 

__Step 1:__ Set-up the data and the variable(s) of interest

__Step 2:__ Define what geometric object should be plotted on the plot

__Step 3:__ Declare nice x- and y-axis labels

Let's carry out these steps in the code chunk below! 

```{r sight-freq-barplot}
ggplot(data = myopia, 
       mapping = aes(x = ____)
       ) + 
  geom_bar() + 
  labs(x = "____", 
       y = "____")
```

__3. Which sight group has the largest number of observations?__

4. Using the bar plot, estimate how many children have some level of myopia.


Using the skills you just acquired, make a bar plot visualizing how many
children in our data set slept with a nightlight (`Light`). 

```{r light-freq-barplot}

```

__5. Which nightlight group has the largest number of observations?__


### Relative Frequency Bar Plots

We could also choose to display the data as a proportion in a
**relative frequency** bar plot. To find the relative frequency, we divide the
count in each level by the total sample size. The resulting values are sample
proportions. 

To get *proportions* instead of *counts* we need to override the default 
statistic `geom_bar()` wants to use. To do this, we tell it we want proportions
on the y-axis. We do this using the following code:

```
geom_bar(aes(y = ..prop.., group = 1))
```

Notice, there are two periods (..) before **and** after prop. These periods are
necessary to define a new statistic that should be used for the y-axis. 

6. Use this new code to modify the code you had previously. Copy-and-paste your
code from #4 below and change the y-axis to plot proportions instead of counts. 

```{r light-prop-barplot}
# Copy-and-paste your code from #4
# Add the aes() input to geom_bar() so your plot has proportions instead of counts!


```


7. Try removing the `group = 1` input from `geom_bar()`. What happens? What do
your bars look like?



## Summary statistics for categorical variables

The other part of Exploratory Data Analysis (EDA) is making summary statistics.
For a categorical variable, this means making frequency tables and relative
frequency tables. Let's explore how to do that. 

For this piece you are going to learn more about a tool you've seen
previously -- "the pipe." This is a special operator written as `%>%`. The pipe
is used as a connecting piece to keep your data flowing through a pipeline. 

We will use the pipe to string together three R functions:

- `summarize()`, which creates summary statistics of specified variables
- `group_by()`, which creates groups of observations based on a categorical 
variable
- `mutate()`, which modifies our dataset by adding new variables or changing
existing variables

If I use `group_by()` in my analysis, it is nearly always paired with 
`summarize()` -- they are function friends!

Let's start with a short data pipeline.

```{r summarize}
myopia %>%
  summarize(n = n())
```

__8. What does the variable n tell us?__


We can add another piece to the pipeline to make it a bit more interesting.
Let's add a `group_by()` step to make groups based on the `Sight` varible. 

```{r group-by-summarize}
myopia %>%
  group_by(___) %>% 
  count()
```

__9. How many observations are in the Myopia and High Myopia groups?__


Finally, we might want to add one more piece to the pipeline to create a 
`proportion` variable of the relative frequencies of each level of `Sight`.

```{r group-by-summarize-mutate}
myopia %>%
  group_by(___) %>% 
  count() %>% 
  mutate(proportion = ___ / sum(n))
```

__10. What proportion of children have high myopia?__


## On Your Own

Now, take the code you just filled in and address the following question about
the `Light` variable!

What proportion of children slept with full light? 


```{r light-proportions}
# Put your code here to find the proportion of children that slept with different nightlights


```

