---
title: "Week 6 Day 1: Confidence Interval for Snowfall between Weather Patterns"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      out.width = "60%")

library(tidyverse)
library(mosaic)
```

### Learning outcomes

* Use bootstrapping to find a confidence interval for a difference in means.

* Interpret a confidence interval for a difference in means.

* Use a confidence interval to determine the conclusion of a hypothesis test.

---

## Review from last week



---

## Confidence interval



---

## Creating a bootstrap distribution in `R`

We will use the `two_mean_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of differences in sample means and calculate a confidence interval. 

19. Using bootstrapping find a 95\% confidence interval. Using the provided `R` script file, enter the variable names and data set name as in the `two_mean_test()` function, outcome name for the first in subtraction, number of repetitions, and the confidence level as a decimal.  Highlight and run lines 32--35. Report the 95\% confidence interval in interval notation.

```{r, echo = TRUE, eval = FALSE}
two_mean_bootstrap_CI(RESPONSE ~ EXPLANATORY, data = DATASET,  # Variables and data
                      first_in_subtraction = "xx", # First value in order of subtraction
                      number_repetitions = 1000,  # Number of simulations
                      confidence_level = xx)
```

\vspace{0.3in}

20. Interpret the interval you calculated in question 19. 

\vspace{1in}


21.  Would the results from a theory-based test match the results we saw with
the simulation?  Explain why or why not.

\vspace{1in}

\newpage

### Take-home messages

1. To create one simulated sample on the bootstrap distribution for a difference
in sample means, label $n_1 + n_2$ cards with the original response values.
Keep groups separate and randomly draw with replacement $n_1$  times from group
1 and $n_2$ times from group 2.  Calculate and plot the resampled difference in
means.  

<!-- Something about theoretical methods! -->