---
title: "STAT 218 - Midterm 1"
date: "October 20, 2022"
author: "Dr. Theobold"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(moderndive)
library(pander)
library(infer)
library(lterdatasampler)
library(palmerpenguins)
```

\begin{center}
\large{Name:} \underline{\hspace{3in}}
\vspace{0.25cm}

\large{Section (circle one):} \hspace{1cm} 2:10pm \hspace{2cm} 4:10pm 

\end{center}

\vspace{0.8cm}

\noindent {\bf Read and Sign the Following Statement:}

I understand that giving or receiving help on this exam is a violation of academic
regulations and is punishable by a grade of __F__ in this course. This
includes looking at other students' exams and / or allowing other students,
actively or passively, to see answers on my exam. This also includes revealing,
actively or passively, any information about the exam to any member of 
Dr. Robinson's STAT 218 class who has not yet taken the exam. The use of 
cell phones is strictly prohibited.

\begin{center}

\textbf{I pledge not to do any of these things.}

\vspace{0.5cm}

\textbf{Signed:} \underline{\hspace{4in}}

\end{center}

\vspace{1cm}

\noindent {\bf \textsf{Instructions}}

\begin{itemize}
\item Read and sign the honesty pledge at the top of this page. Your exam will
not be graded unless the honesty pledge is signed!
\item You may use a calculator. You \textbf{may not} use your phone or any
device that connects to the internet as a calculator. \\
\item Show all work as clearly as possible. Point totals are shown in brackets 
next to each part. Formulas without values entered do not count as work. \\
\item All answers should be reported in decimal form, rounded to three decimal
places. \\
\item For multiple choice and multi-select problems, completely fill in the
provided circle (multiple choice) or square (multi-select) for your desired
answer choice(s). If you change an answer, be sure to completely erase your
initial selection.
\item You have 50 minutes to complete this exam, so budget your time wisely.    
\end{itemize}

\pagebreak

\noindent {\bf \textsf{Provided Formulas}}

$$R^2 = r^2$$ 

\vspace{0.25cm}

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

$$SE(\bar{x}) = \frac{s}{\sqrt{n}}$$
\vspace{0.3cm}

__general formula for a confidence interval:__ $\text{point estimate} \pm \text{multiplier} \times \text{SE(point estimate)}$

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\bar{x} \pm t^*_{df} \times SE(\bar{x})$

\newpage


\vspace{0.25cm}

\noindent \textbf{Q8} [2 points] When you change from a 90% to a 95% confidence
interval, which part(s) of the confidence interval change? (Select all that 
apply)

\begin{itemize}
\item[(a)] Point estimate (midpoint)
\item[(b)] Multiplier
\item[(c)] Standard error
\end{itemize}

\vspace{0.75cm}

\noindent \textbf{Q6} [3 points] The purpose of creating a null distribution is
to: (Select all that apply)

\begin{itemize}
\item[(a)] Discover what statistics might have occurred if the null hypothesis
was true.
\item[(b)] To determine if the null hypothesis is true. 
\item[(c)] To determine if the observed statistic is unlikely if the null was
true. 
\end{itemize}

\vspace{0.75cm}

\noindent \textbf{Q5} [4 points] Indicate whether each statement about a 
bootstrap resample is \textbf{TRUE} or \textbf{FALSE}.

\begin{itemize}
\item[(a)] The bootstrap resample and original sample \textbf{must} be the same size.
\hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(b)] The bootstrap resample and original sample are \textbf{both} taken
directly from the population. 

\item[] \hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(c)] The bootstrap resample can \textbf{only} use values that were in the
original sample. \hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(d)] The bootstrap resample uses \textbf{all} of the values that were in
the original sample. \hspace{0.1cm} \underline{\hspace{2cm}} \\
\end{itemize}

\newpage

\textbf{Q3} The Atlantic marsh Fiddler Crab, *Minuca pugnax*, lives in salt
marshes throughout the eastern coast of the United 
States. Historically, M. pugnax were distributed from northern Florida to Cape
Cod, Massachusetts, but like other species have expanded their range northward
due to ocean warming.

The Plum Island Ecosystem Long Term Ecological Research site collected data on
Fiddler Crabs from 13 marshes on the Atlantic coast of the United States in the
summer of 2016. The marshes spanned from northeast Florida to northeast
Massachusetts. Researchers were able to collect between 25 and 37 adult male
Fiddler Crabs at each marsh. 

\vspace{0.2cm}

\noindent 
\textbf{(a)} [3 pts] A preview of the dataset is provided below. Use this
preview to address the following questions. 

```{r crab-preview}
pie_crab |>  
  select(-date, -air_temp_sd, -water_temp_sd)
```

\begin{itemize}
\item Identify the observational units in the data set. \\

\vspace{1cm}

\item List the variables. Indicate whether each variable is categorical (c) or
quantitative (q). \\

\vspace{2cm}

\end{itemize}

\textbf{(b)} [3 pts] A histogram displaying the size of the sample of Fiddler
Crabs is displayed below. Describe the shape of the distribution. Be sure to 
address the center, spread, shape, and outliers. 

```{r, fig.align = 'right', out.width = "30%"}
pie_crab %>% 
  ggplot(aes(x = size)) +
  geom_histogram(binwidth = 1.5, color = "white") +
  labs(x = "Carapace Width (millimeters)", 
       y = "Number of Crabs")
```

\vspace{0.25cm}

\textbf{(c)} [2 pts] The researchers collected data on 392 total Fiddler Crabs. 
When using a $t$-distribution to find a 95% confidence interval for $\mu$, how
many degrees of freedom should be used?

\vspace{1cm}

```{r}
crab_ci <- t_test(pie_crab, 
       response = size,
       conf_int = TRUE, 
       conf_level = 0.95) %>% 
  select(estimate, lower_ci, upper_ci)
```

\textbf{(d)} [4 pts] A 95% confidence interval for the mean carapace width of
Fiddler Crabs was found to be (`r round(crab_ci$lower_ci, 2)`, 
`r round(crab_ci$upper_ci, 2)`). Below is the researchers' interpretation of
this confidence interval: 

\begin{center}
\emph{There is a 95\% chance that the observed sample mean carapace width of the
392 Fiddler Crabs is between 14.31 and 15.01 millimeters.}
\end{center}

\noindent
Identify __three__ mistakes committed and fix them. Be brief but clear in your
description.

\vspace{1cm}

Mistake 1: \underline{\hspace{6in}}

\vspace{1cm}

Fix: \underline{\hspace{6.2in}}

\vspace{1.5cm}

Mistake 2: \underline{\hspace{6in}}

\vspace{1cm}

Fix: \underline{\hspace{6in}}

\vspace{1.5cm}

Mistake 3: \underline{\hspace{6in}}

\vspace{1cm}

Fix: \underline{\hspace{6in}}

\vspace{0.25in}

\textbf{(e)} [2 points] Can the researchers use their interval to make
inferences about __all__ Fiddler Crabs in the United States? Justify your 
answer!

\newpage

\noindent {\textbf Q2} I collected data on 512 different fast food items from
Mcdonalds, Chick-Fil-A, Sonic, Arby's, Burger King, Dairy Queen, Subway, and
Taco Bell. To obtain these data, for every restaurant I randomly sampled 64
items from their entire menu and recorded the nutritional content of each item
selected (e.g., calories, saturated fat, calcium, protein, etc.). 

\vspace{1cm}

\noindent
\textbf{(a)} [2 pts] Describe the sampling method I used to obtain these 512 
fast food items. 

\vspace{1.5in}

\noindent
\textbf{(b)} [3 pts] I am interested in studying the linear relationship between
the total calories of a food item and the amount of saturated fat that item
contains.

\textbf{Write the null hypothesis for my question of interest, using both 
\underline{words} and \underline{notation}.}

\vspace{1in}

\noindent
\textbf{(c)} [2 pts] Is the alternative hypothesis one-sided or two-sided?
(Circle one.)

\begin{itemize} 
\item One-sided \\
\item Two-sided
\end{itemize}

\noindent
\textbf{(d)} [5 pts] Below is the plot of the simulated null
distribution from `R`. Fill in the blanks below to correctly explain how one
sample on the null distribution was created.  Blanks preceded by (#) should
be filled in with a number, all other blanks should be filled in with either the
context of the study or the procedure that would need to be carried out to 
obtain one simulated sample.

\begin{center}
On (\#) \underline{\hspace{2in}} cards, write \underline{\hspace{1in}} and 
\underline{\hspace{1in}} on the cards.

\vspace{0.5cm}

Assume the null hypothesis is true and \underline{\hspace{3in}}.

\vspace{0.5cm}

Generate a new sample of 512 ordered pairs by \underline{\hspace{3in}}.

\vspace{0.5cm}

Calculate and plot the \underline{\hspace{4cm}} from each simulated sample.

\end{center}

```{r null-dist-fastfood, fig.height = 3.5}

obs_slope <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  calculate(stat = "slope") %>% 
  pull()
  
null_dist <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, 
           type = "permute") %>% 
  calculate(stat = "slope")

visualise(null_dist) +
  labs(x = "Simulated Slope Statistic", 
       y = "Frequency (out of 1000 reps)", 
       title = "") +
  xlim(c(-0.02, 0.02))
```

\noindent
\textbf{(e)} [2 pts] Using the regression output below, draw a vertical line
where the observed statistic falls on the null distribution. 

```{r}
fastfood_lm <- lm(sat_fat ~ calories, data = fastfood)

get_regression_table(fastfood_lm) %>% 
  select(term, estimate, std_error) %>% 
  pander()
```

\vspace{0.25cm}

\noindent
\textbf{(f)} [2 pts] Shade the location of the plot you would use to calculate
the p-value. 

\vspace{0.25cm}

\noindent
\textbf{(g)} [1 pts] Estimate the p-value associated with this hypothesis test. 

\vspace{0.5cm}

\noindent
\textbf{(h)} [3 pts] Which of the following is a correct interpretation of the p-value obtained? (Circle one)

(a) If there is no linear relationship between the total calories and the
saturated fat of a fast food item, we would observe a sample slope of 0.017 or
more extreme with a probability of less than 1 in 1000. 

(b) If there is a linear relationship between the total calories and the
saturated fat of a fast food, we would observe a sample slope of 0.017 or more
extreme with a probability of less than 1 out of 1000.

(c) The probability of seeing a sample slope between the total calories and
the saturated fat of a fast food item of 0.017 or more extreme is less than 0.1%.

(d)  The probability that there is no linear relationship between the total
calories and the saturated fat of a fast food item, is less than 0.1%.

\vspace{0.5cm}

\textbf{(i)} [2 points] Given the p-value for the hypothesis test, would the
95% confidence interval for $\beta_1$ contain 0? Be sure to justify your 
choice!

\newpage

\noindent \textbf{Q11} The Konza Prairie Long-Term Ecological Research has
collected data on bison on the Konza prairie since 1994, making it the longest
continuous record of wild ungulate weight gain anywhere in the world.
Researchers conduct a round-up once a year at the end of the grazing season
wherein each bison is weighed, calves are vaccinated and receive unique IDs, and
excess individuals are culled.

For this investigation, we are interested in assessing if, despite the effect 
of climate change on their habitat, the weight of yearling, male bison is what
is described as "healthy" --- a weight of approximately 750 pounds. 

Below are summary statistics for the 48 of the yearling, male bison captured
in 2020. 

```{r bison-stats}
yearling_male_bison <- knz_bison %>% 
  mutate(age = rec_year - animal_yob) %>% 
  group_by(animal_code) %>% 
  filter(age == 1, 
         animal_sex == "M", 
         rec_year == 2020)

favstats(~ animal_weight, data = yearling_male_bison) %>% 
  pander()
```


\textbf{(a)} [3 points] Define the parameter of interest in words and use proper notation to assign a symbol.

\vspace{0.75in}

\textbf{(c)} [2 points] If we decided to use inferential methods to assess if the
true mean weight of yearling bison on the Konza Prairie was healthy, we must
verify two conditions. What are these two conditions? 

\vspace{0.25cm}

Condition 1:

\vspace{1in}

Condition 2: 

\vspace{1in}

\begin{center}

\textbf{END OF EXAM}

\end{center}

---

\vspace{1in}

__Points Earned:__ \underline{\hspace{2in}}

__Total Points: 46__

