---
title: "Midterm 1 Question Bank"
author: "Stat 218"
output: 
  pdf_document:
    extra_dependencies: ["multicol", "setspace"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(moderndive)
library(pander)
library(infer)
library(lterdatasampler)
library(palmerpenguins)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted, so it is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours.

\vspace{0.5cm}

## Provided Formulas

$$R^2 = r^2$$ 

\vspace{0.25cm}

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

$$SE(\bar{x}) = \frac{s}{\sqrt{n}}$$
\vspace{0.3cm}

__general formula for a confidence interval:__ $\text{point estimate} \pm \text{multiplier} \times \text{SE(point estimate)}$

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\bar{x} \pm t^*_{df} \times SE(\bar{x})$

\newpage

\noindent {\textbf Q1} Dr. John Arbuthnot, an 18th century physician, writer, 
and mathematician is famous for performing the first hypothesis test of
significance. Dr. Arbuthnot was interested in the ratio of newborn males to
newborn females, so he gathered the baptism records for children born in London
for every year from 1629 to 1710. Artbuthnot found that in every year, the
number of males born in London exceeded the number of females.

\noindent
\textbf{(a)} [2 pts] Describe the sampling method used by Dr. Arbuthnot.

\vspace{0.75in}

\noindent 
\textbf{(b)} [2 pts] Describe how this sampling method could be biased.

\vspace{0.75in}

\noindent 
\textbf{(c)} [3 pts] A preview of the dataset is provided below. Use this
preview to address the following questions. 

```{r data-preview}
arbuthnot
```

\begin{itemize}
\item Identify the observational units in the data set. \\

\vspace{1cm}

\item List the variables. Indicate whether each variable is categorical or
quantitative. \\

\vspace{1cm}

\item What would the dimensions of the data set be? (number of rows by number of
columns)

\end{itemize}

\vspace{2cm}

\newpage

\textbf{(d)} [3 pts] A scatterplot displaying the number of girls born over time
is displayed below. Describe the relationship you see in the scatterplot. Be
sure to address the form, direction, strength, and outliers present.

```{r arbuthnot-scatter}
ggplot(data = arbuthnot, 
       aes(x = year, y = girls)) + 
  geom_point() + 
  labs(x = "Year", 
       y = "Number of Girls")
```

\vspace{2cm}

\textbf{(e)} [2 pts] Would it be appropriate to model the relationship between
the number of girls born and the year with a linear regression? Justify your
belief! 

\newpage

\noindent {\textbf Q2} I collected data on 512 different fast food items from
Mcdonalds, Chick-Fil-A, Sonic, Arby's, Burger King, Dairy Queen, Subway, and
Taco Bell. To obtain these data, for every restaurant I randomly sampled 64
items from their entire menu and recorded the nutritional content of each item
selected (e.g., calories, saturated fat, calcium, protein, etc.). 

\noindent
\textbf{(a)} [2 pts] Describe the sampling method I used to obtain these 512 
fastfood items. 

\vspace{2cm}

\noindent
\textbf{(b)} [3 pts] I am interested in studying the linear relationship between
the total calories of a food item and the amount of saturated fat that item
contains.

__Write the null hypothesis for my question of interest, using both words and
notation.__

\vspace{2cm}

\noindent
\textbf{(c)} [2 pts] Is the alternative hypothesis one- or two-sided? Select
one.

\begin{itemize} 
\item One-sided \\
\item Two-sided
\end{itemize}

\noindent
\textbf{(d)} [5 pts] On the following page is the plot of the simulated null
distribution from `R`. Fill in the blanks below with one answer in each set of
parentheses to correctly explain how one sample on the null distribution would
be created.  Blanks preceded by (#) should be filled in with a number, all
other blanks should be filled in with the context of the study.

\begin{center}
On (\#) \underline{\hspace{2in}} cards, write \underline{\hspace{1in}} and 
\underline{\hspace{1in}} on the cards.

\vspace{0.5cm}

Assume the null hypothesis is true and \underline{\hspace{4in}}.

\vspace{0.5cm}

Generate a new sample of 512 ordered pairs by \underline{\hspace{3.5in}}.

\vspace{0.5cm}

Calculate and plot the \underline{\hspace{4cm}} from each simulated sample.

\end{center}

```{r null-dist-fastfood}

obs_slope <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  calculate(stat = "slope") %>% 
  pull()
  
null_dist <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, 
           type = "permute") %>% 
  calculate(stat = "slope")

visualise(null_dist) +
  labs(x = "Simulated Slope Statistic", 
       y = "Frequency (out of 1000 reps)", 
       title = "") +
  xlim(c(-0.02, 0.02))
```

\noindent
\textbf{(e)} [2 pts] Using the regression output below, draw a vertical line
where the observed statistic falls on the null distribution. 

```{r}
fastfood_lm <- lm(sat_fat ~ calories, data = fastfood)

get_regression_table(fastfood_lm) |> 
  select(term, estimate, std_error) |> 
  pander() 
```

\vspace{0.25cm}

\noindent
\textbf{(f)} [2 pts] Shade the location of the plot you would use to calculate
the p-value. 

\vspace{0.25cm}

\noindent
\textbf{(g)} [1 pts] Estimate the p-value associated with this hypothesis test. 

\vspace{0.5cm}

\noindent
\textbf{(h)} [3 pts] Which of the following is a correct interpretation of the
p-value obtained? (Circle one)

\begin{itemize}
\item If there is no linear relationship between the total calories and the
saturated fat of a fast food item, we would observe a sample slope of 0.017 or
more extreme with a probability of less than 1 in 1000. 

\item If there is a linear relationship between the total calories and the
saturated fat of a fast food, we would observe a sample slope of 0.017 or more
extreme with a probability of less than 1 in 1000.

\item The probability of seeing a sample slope between the total calories and
the saturated fat of a fast food item of 0.017 or more extreme is less than 0.1%. 
\item  The probability that there is no linear relationship between the total
calories and the saturated fat of a fast food item, is less than 0.1%.
\end{itemize}

\textbf{(i)} [2 points] Given the p-value for the hypothesis test, would the
95% confidence interval for $\beta_1$ contain 0? Be sure to justify your choice!

\newpage

\textbf{Q3} The Atlantic marsh Fiddler Crab, *Minuca pugnax*, lives in salt
marshes throughout the eastern coast of the United 
States. Historically, M. pugnax were distributed from northern Florida to Cape
Cod, Massachusetts, but like other species have expanded their range northward
due to ocean warming.

The Plum Island Ecosystem Long Term Ecological Research site collected data on
Fiddler Crabs from 13 marshes on the Atlantic coast of the United States in the
summer of 2016. The marshes spanned from northeast Florida to northeast
Massachusetts. Researchers were able to collect between 25 and 37 adult male
fiddler crabs at each marsh. 

\noindent 
\textbf{(a)} [3 pts] A preview of the dataset is provided below. Use this
preview to address the following questions. 

```{r crab-preview}
pie_crab %>% 
  select(-name, -air_temp_sd, -water_temp_sd) 
```

\begin{itemize}
\item Identify the observational units in the data set. \\

\vspace{1cm}

\item List the variables. Indicate whether each variable is categorical or
quantitative. \\

\vspace{1cm}

\end{itemize}

\textbf{(b)} [3 pts] A histogram displaying the size of the sample of fiddler
crabs is displayed below. Describe the shape of the distribution. Be sure to 
address the center, spread, shape, and outliers. 

```{r}
pie_crab %>% 
  ggplot(aes(x = size)) +
  geom_histogram(binwidth = 1.5, color = "white") +
  labs(x = "Carapace Width of Fiddler Crab (millimeters)", 
       y = "Number of Crabs")
```

\vspace{0.25cm}

\textbf{(c)} [2 pts] The researchers collected data on 392 total Fiddler Crabs. 
When using a $t$-distribution to find a 95% confidence interval for $\mu$, how
many degrees of freedom should be used?

\vspace{1cm}

```{r}
crab_ci <- t_test(pie_crab, 
       response = size,
       conf_int = TRUE, 
       conf_level = 0.95) %>% 
  select(estimate, lower_ci, upper_ci)
```

\textbf{(d)} [6 pts] A 95% confidence interval for the mean carapace width of
Fiddler Crabs was found to be (`r round(crab_ci$lower_ci, 2)`, 
`r round(crab_ci$upper_ci, 2)`). Below is the researchers' interpretation of
this confidence interval: 

\begin{center}
\emph{There is a 95\% chance that that the observed sample mean of the 392 
Fiddler Crabs sampled will have a mean carapace width between 14.31 and 15.01
millimeters.}
\end{center}

\noindent
Identify __three__ mistakes committed and fix them. Be brief but clear in your
description.

\vspace{.15in}

Mistake 1: \underline{\hspace{6in}}

\vspace{0.5cm}

Fix: \underline{\hspace{6in}}

\vspace{0.5cm}

Mistake 2: \underline{\hspace{6in}}

\vspace{0.5cm}

Fix: \underline{\hspace{6in}}

\vspace{0.5cm}

Mistake 3: \underline{\hspace{6in}}

\vspace{0.5cm}

Fix: \underline{\hspace{6in}}

\newpage

\textbf{(e)} [2 points] Can the researchers use their interval to make
inferences about __all__ Fiddler Crabs in the United States? Justify your 
answer!

\vspace{1in}

\noindent \textbf{Q4} [4 points] Researchers are interested in the fish that
reside in the Caspian Sea. They have plans to collect many fish and take
multiple measurements on each. Match each statistical description on the right
with each piece of information given. Put the letter of the statistical
description in the blanks on the left.

\begin{center}
	\begin{tabular}{p{8cm}p{6cm}}
		\underline{\hspace{.4in}} circumference of the fish &	(a) quantitative variable \\
		&   \\
		\underline{\hspace{.4in}} species of the fish																																															&	(b) categorical variable \\
		&   \\
		\underline{\hspace{.4in}} average length of all fish in \\ the area of consideration																													&		(c) parameter: $\mu$	\\
		&   \\
		\underline{\hspace{.5in}} mean internal temperature \\ of the fish collected in the sample																										&	(d) statistic: $\bar{x}$	\\
		&   \\
		\underline{\hspace{.4in}} one of the fish in the area of \\ consideration																																		&	(e) observational unit	\\
		&   \\
		\underline{\hspace{.4in}} method of only studying the fish \\ caught in the net \@ 3pm on Wednesday of the research time frame								&	(f) cluster sampling method	\\
		&   \\
		\underline{\hspace{.4in}} method of selecting 5\% of each \\ species, known to be in the area of consideration, for the sample								&	(g) stratified sampling method	\\
		&   \\
		\underline{\hspace{.3in}} method of dividing up the whole \\ location with netting and sampling 10 random netted areas												& (h) convenience sampling method
	\end{tabular}
\end{center}

\vspace{0.5cm}

\noindent \textbf{Q5} [4 points] Indicate whether each statement about a 
bootstrap resample is \textbf{TRUE} or \textbf{FALSE}.

\begin{itemize}
\item[(a)] The bootstrap resample and original sample \textbf{must} be the same size.
\hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(b)] The bootstrap resample and original sample are \textbf{both} taken
directly from the population. 

\item[] \hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(c)] The bootstrap resample can \textbf{only} use values that were in the
original sample. \hspace{0.1cm} \underline{\hspace{2cm}} \\

\item[(d)] The bootstrap resample uses \textbf{all} of the values that were in
the original sample. \hspace{0.1cm} \underline{\hspace{2cm}} \\
\end{itemize}

\newpage

\noindent \textbf{Q6} [3 points] The purpose of creating a null distribution is
to: (Select all that apply)

\begin{itemize}
\item[(a)] Discover what statistics might have occurred if the null hypothesis
was true.
\item[(b)] To determine if the null hypothesis is true. 
\item[(c)] To determine if the observed statistic is unlikely if the null was
true. 
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{Q7} [2 points] An article in the San Luis Tribune claims that
the average age for people who receive food stamps in SLO is 40 years. A Cal
Poly student believes the average age is less than that. The student obtains a
random sample of 100 people in SLO who receive food stamps, and finds their
average age to be 39.2 years. Performing a hypothesis test, the student finds 
their sample mean to be statistically significantly lower than the age of 40 
stated in the article (p-value < 0.05). Indicate fore each of the following
interpretations whether they are valid or invalid.

\noindent \textbf{(a)} The statistically significant result indicates that the 
majority of people who receive food stamps is younger than 40.
\begin{itemize}
\item Valid
\item Invalid
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{(b)} An error must have been made. This difference in means
(39.2 vs. 40 years) is too small to be statistically significant.
\begin{itemize}
\item Valid
\item Invalid
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{Q8} [2 points] When you change from a 90% to a 95% confidence
interval,which part(s) of the confidence interval change? (Select all that 
apply)

\begin{itemize}
\item[(a)] Statistic (midpoint)
\item[(b)] Multiplier
\item[(c)] Standard error
\end{itemize}

\vspace{0.25cm}

\noindent \textbf{Q9} When using a $t$-distribution to obtain a confidence
interval, how does the multiplier change from the 95% to the 90% confidence
interval? (Circle the correct answer)

\begin{itemize}
\item[(a)] Multiplier is larger
\item[(b)] Multiplier is smaller
\item[(c)] Multiplier stays the same
\end{itemize}

\newpage

\noindent \textbf{Q10} Based on the plots below, which is the better measure of
center for this variable? (Circle one.)

\begin{itemize}
\item[(a)] The median, as the shape of the distribution is skewed.
\item[(b)] The mean, as the shape of the distribution is symmetric.
\item[(c)] The mean, as the shape of the distribution is skewed.
\item[(d)] The median, as the shape of the distribution is symmetric.
\end{itemize}

```{r, fig.width = 8}
hist <- penguins %>% 
  ggplot(mapping = aes(x = body_mass_g)
         ) +
  geom_histogram(color = "white", binwidth = 150) + 
  labs(x = "Body Mass (grams)")

boxplot <- penguins %>% 
  ggplot(mapping = aes(x = body_mass_g)
         ) +
  geom_boxplot() + 
  labs(x = "Body Mass (grams)") + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())

gridExtra::grid.arrange(hist, boxplot, nrow = 1)
```

\vspace{0.5cm}

```{r bison-filter}
yearling_male_bison <- knz_bison %>% 
  mutate(age = rec_year - animal_yob) %>% 
  group_by(animal_code) %>% 
  filter(age == 1, 
         animal_sex == "M", 
         rec_year == 2020)

```

\noindent \textbf{Q11} The Konza Prairie Long-Term Ecological Research has
collected data on bison on the Konza prairie since 1994, making it the longest
continuous record of wild ungulate weight gain anywhere in the world.
Researchers conduct a round-up once a year at the end of the grazing season
wherein each bison is weighed, calves are vaccinated and receive unique IDs, and
excess individuals are culled.

For this investigation, we are interested in assessing if, despite the effect 
of climate change on their habitat, the weight of yearling, male bison is what
is described as "healthy" --- a weight of approximately 750 pounds. 

Below are summary statistics for the 48 of the yearling, male bison captured
in 2020. 

```{r bison-stats}
favstats(~ animal_weight, data = yearling_male_bison) %>% 
  pander()
```


\textbf{(a)} [3 points] Define the parameter of interest, using proper notation.

\pagebreak

\textbf{(b)} [2 points] What type of inference should be done to answer the
research question? (Circle one)

\begin{itemize}
\item Confidence Interval
\item Hypothesis Test
\item No inference is needed. The true average weight of a yearling male bison
is 629.5 pounds.
\item No inference is needed.  The true average weight of a yearling male bison
is 750 pounds.
\end{itemize}

\vspace{0.25cm}

\textbf{(c)} [2 points] If we decided to use inferential methods to assess if the
true mean weight of yearling bison on the Konza Prairie was healthy, we must
verify two conditions. What are these two conditions? 

\vspace{0.25cm}

Condition 1:

\vspace{1in}

Condition 2: 

\vspace{1in}

\textbf{(d)} [2 points] Using the plot below and the description of the data
collection procedure, justify if you believe whether each condition is 
violated or not. 

\vspace{0.25cm}

Condition 1: 

\vspace{1in}

Condition 2: 

\vspace{1in}

```{r bison-hist}
yearling_male_bison |>  
  ggplot(mapping = aes(x = animal_weight)
         ) +
  geom_histogram(binwidth = 30, color = "white") +
  labs(x = "Bison Weight (lbs)", 
       y = "Number of Bison")
```
