---
title: "Final Exam Question Bank"
author: "Stat 218"
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - multicol
    - setspace
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(moderndive)
library(pander)
library(infer)
#library(lterdatasampler)
library(palmerpenguins)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted. It is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours prior to the exam.

\vspace{0.5cm}

## Golden Ticket

| Scenario | One Categorical Response | Two Categorical Variables | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Bar plot | Side-by-Side Bar plot, Stacked Bar plot, Mosaic plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
| Summary measure | Proportion | Difference in Proportions | Mean | Slope or Correlation | Difference in Means |
| Parameter notation | $\pi$ | $\pi_1 - \pi_2$ | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
| Statistic notation | $\widehat{p}$ | $\widehat{p}_1 - \widehat{p}_2$ | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\mu_1 - \mu_2) = \sqrt{\frac{\sigma_1}{n_1} + \frac{\sigma_2}{n_2}}$$

\vspace{0.25cm}

$$\widehat{p}_{pool} = \frac{\text{total number of "successes"}}{\text{total number of cases}} = \frac{\hat{p}_1 * n_1 + \hat{p}_2 * n_2}{n_1 + n_2}$$ 

\newpage 

\noindent
\textbf{Q1}[]
Wild mushrooms, such as chanterelles or morels, are delicious, but eating wild
mushrooms carries the risk of accidental poisoning. Even a single bite of the
wrong mushroom can be enough to cause fatal poisoning. An amateur mushroom
hunter is interested in finding an easy rule to differentiate poisonous and
edible mushrooms. They think that the mushroom’s gills (the part which holds and
releases spores) might be related to a mushroom’s edibility. They used a data
set of 8124 mushrooms and their descriptions. For each mushroom, the data set
includes whether it is edible or poisonous and the spacing of the gills (Broad
or Narrow). 

**Please Note**: According to The Audubon Society Field Guide to North American
Mushrooms, there is no simple rule for determining the edibility of a mushroom;
no rule like "leaflets three, leave them be" for Poisonous Oak and Ivy.

|       | Broad Gill Size | Narrow Gill Size | Total |
|-------|-----------------|------------------|-------|
| Edible | 3920 | 288 | 4208 |
| Poisonous | 1692 | 2224 | 3916 |
| Total | 5612 | 2512 | 8124 |

\vspace{0.2cm}

\noindent
\textbf{(a)}[4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.2cm}

Whether the mushroom is edible or poisonous is the (explanatory / response) \underline{\hspace{3in}} and it is
(categorical / quantitative) \underline{\hspace{3in}}. 

\vspace{0.2cm}

Gill size (Broad or Narrow) is the (explanatory / response) 
\underline{\hspace{3in}} and it is (categorical / quantitative)
\underline{\hspace{3in}}. 

\vspace{0.2cm} 

\noindent
\textbf{(b)}[3 points] Calculate the proportion of mushrooms with a broad gill
size that are poisonous. *Give appropriate notation with informative subscripts
if necessary.* 

\vspace{0.2cm}

**Work:**

\vspace{1cm}

**Answer:** \underline{\hspace{3in}}

\vspace{0.2cm}

**Notation:** \underline{\hspace{3in}}

\vspace{0.2cm}

\noindent
\textbf{(c)}[3 points] Calculate the proportion of mushrooms with a narrow gill
size that are poisonous. *Give appropriate notation with informative subscripts
if necessary.* 

\vspace{0.2cm}

**Work:**

\vspace{1cm}

**Answer:** \underline{\hspace{3in}}

\vspace{0.2cm}

**Notation:** \underline{\hspace{3in}}

\vspace{0.2cm}

\noindent
\textbf{(d)}[3 points] Using your answers to (b) and (c), fill in the blanks
below to label the bar chart showing the relationship between gill size (broad
or narrow) and whether the mushroom is edible. 

```{r mushroom-barplot}

## Make a barplot
## Add blanks where the legend is so students fill in!

```


