---
title: "STAT 218 - Midterm Exam 2"
date: "November 10, 2020"
author: "Dr. Theobold"
output:
  pdf_document:
    extra_dependencies: ["multicol", "setspace"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(pander)
library(infer)
library(broom)
```

\begin{center}

\large{Name:} \underline{\hspace{3in}}
\vspace{0.25cm}

\end{center}

\vspace{0.8cm}

\noindent {\bf Read and Sign the Following Statement:}

I understand that give or receiving help on this exam is a violation of academic
regulations and is punishable by a grade of __F__ in this course. This
includes looking at other students' exams and / or allowing other students,
actively or passively, to see answers on my exam. This also includes revealing,
actively or passively, any information about the exam to any member of 
Professor Theobold's STAT 218 class who has not yet taken the exam. The use of 
cell phones is strictly prohibited.

\begin{center}

\textbf{I pledge not to do any of these things.}

\vspace{0.5cm}

\textbf{Signed:} \underline{\hspace{4in}}

\end{center}

\vspace{1cm}

\noindent {\bf \textsf{Instructions.}}

\begin{itemize}
\item Read and sign the honesty pledge at the top of this page. Your exam will
not be graded unless the honesty pledge is signed!
\item You may use a calculator. You \textbf{may not} use your phone or any
device that connects to the internet as a calculator. \\
\item Show all work as clearly as possible. Point totals are shown in brackets 
next to each part. Formulas without values entered do not count as work. \\
\item All answers should be reported in decimal form, rounded to three decimal
places. \\
\item For multiple choice and multi-select problems, circle your desired
answer choice(s). If you change an answer, be sure to completely erase your
initial selection.
\item You have 50 minutes to complete this exam, so budget your time wisely.    
\end{itemize}

\pagebreak

## Golden Ticket

*Note you may need to think critically about how to extend some of this information to specific situations not listed.*

| Scenario | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
|           |           |           |
| Summary measure | Mean or Mean Difference | Slope or Correlation | Difference in Means |
|           |           |           |
| Parameter notation | $\mu$ or $\mu_d$  | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
|           |           |           |
| Statistic notation | $\bar{x}$ or $\bar{x}_d$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x \hspace{5cm} Residual = y - \hat{y}$$

\vspace{0.25cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\bar x) = \frac{s}{\sqrt{n}} \hspace{5cm} SE(\bar x_1 - \bar x_2) = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$$

\vspace{0.25cm}

$$F = \frac{\text{MSG}}{\text{MSE}} \hspace{5cm} \alpha^* = \frac{\alpha}{\text{\# of comparisons}}$$


\newpage 

\textbf{Q1} [10 points] Researchers in Southern England collected data on
grassland butterflies. They
were interested in whether movement patterns varied across species and between
male and female butterflies. Researchers observed 164 butterflies over the
three-year length of this study, of which 28 were female and 136 were male. 
These 164 butterflies were considered to be representative of all grassland
butterflies. The butterfly movements were observed by measuring how far they
flew (in meters) from one landing site to the next, called step distance. This
was done by placing a flag at each landing site and measuring the distance
between the flags using a mapping software.

The researchers were interested in investigating if there was a difference
between how far male and female butterflies travel, on average, between landing
sites. 

\vspace{0.5cm}

\noindent
\textbf{(a)} [4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.25cm}

Step distance is the (explanatory/response) \underline{\hspace{3in}} and it is (categorical/quantitative) \underline{\hspace{3in}}. 

\vspace{0.25cm} 

Sex is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}.

\vspace{0.5cm}

\noindent
\textbf{(b)} [2 pts] Which visualizations would be most appropriate to display
the relationship between step distance and sex of the butterfly? Select all that 
apply. 

\vspace{0.15cm}

\begin{itemize}
\item[(i)] Segmented Bar plot \\
\item[(ii)] Scatterplot \\
\item[(iii)] Side-by-side Boxplot \\
\item[(iv)] Faceted histograms 
\end{itemize}

\vspace{0.5cm}

\noindent
\textbf{(c)} [3 pts] Assuming a statistical difference in step distance is
found between the male and female butterflies in the sample, what is the scope
of inference for this study? Select one. 

\vspace{0.15cm}

\begin{itemize}
\item[(i)] Sex causes a difference in average step distance for all grassland
butterflies. \\
\item[(ii)] Sex is associated with a difference in average step distance for
the sample of grassland butterflies. \\
\item[(iii)] Sex causes a difference in average step distance for the sample
of grassland butterflies. \\
\item[(iv)] Sex is associated with a difference in average step distance for all
grassland butterflies. 
\end{itemize}

\newpage 

\noindent
\textbf{Q5} [20 points] The *Journal of Food and Agriculture* contained an
article titled "Influence of hydroponic ad soil cultivation on quality and shelf
life of ready-to-eat lamb's lettuce." In this article, researchers studied the
effects of different hydroponic growing methods on the nitrate content of
lettuce. In their study, the researchers randomly assigned 34 lettuce seedlings
to one of three growing conditions: soil, hydroponic A, or hydroponic B. At the
end of the growing period (60 days), nitrate measurements of the lettuce were
taken (mg / kg). 

```{r lettuce-sim-data}
lettuce <- tibble(
  trt_group = c(
    rep("soil", 9),
    rep("hydro_A", 12),
    rep("hydro_B", 13)
  ), 
  growth = c(
    rnorm(n = 9, mean = 3806, sd = 124), 
    rnorm(n = 12, mean = 4695, sd = 107), 
    rnorm(n = 13, mean = 3878, sd = 103)
    )
)
```

Results from the study are presented in the table below. 

```{r lettuce-table}
lettuce %>% 
  mutate(trt_group = fct_recode(trt_group, 
                                 "Hydroponic A" = "hydro_A", 
                                 "Hydroponic B" = "hydro_B", 
                                 Soil = "soil"
                                 ), 
         trt_group = fct_relevel(trt_group, 
                                 "Soil")
         ) %>% 
    group_by(trt_group) %>% 
  summarize(mean = mean(growth), 
            sd = sd(growth), 
            n = n()
            ) %>% 
  rename(`Treatment Group` = trt_group, 
         `Mean Growth` = mean, 
         `Standard Deviation of Growth` = sd, 
         `Sample Size` = n) %>% 
  pander()
```


\noindent
\textbf{(a)} [2 pts] One of the researcher's main questions was to determine
whether the growing method affects nitrate concentration in lettuce. Considering
how this study was executed, can they address this question? *Briefly justify
your answer.*

\vspace{1in}

Below is an incomplete ANOVA table, summarizing the data. You may use this 
information for the subsequent problems. 

```{r}
aov(growth ~ trt_group, data = lettuce) %>% 
  tidy() %>% 
  mutate(term = fct_recode(term, 
                           "Growing Method" = "trt_group"
                           ), 
         df = c("___", 31), 
         statistic = c("___", NA),
         p.value = scales::pvalue(p.value, accuracy = 0.0001)) %>%
  pander()
  
```

\vspace{0.5cm}

\noindent
\textbf{(b)} [3 pts] In the context of the research and in plain English, 
what are the null and alternative hypotheses investigated in the ANOVA table 
above?

\vspace{0.25cm}
\noindent \textbf{$H_0$:}

\vspace{0.35in}

\noindent \textbf{$H_A$:}

\vspace{0.5in}

\noindent
\textbf{(c)} [2 pts] Rewrite the null hypothesis above to use **notation** for the
parameters that are being tested. 

\vspace{0.25cm}

\textbf{$H_0$:}

\vspace{0.25in}

\textbf{$H_A$:}

\newpage

\noindent
\textbf{(d)} [2 pts] The alternative hypothesis investigated in the ANOVA 
table above is $H_A$: $\mu_{\text{Soil}} \neq \mu_{\text{Hydroponic A}} \neq 
\mu_{\text{Hydroponic B}}$. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(e)} [1 pts] What are the degrees of freedom associated with 
`Growing Method`? 

\vspace{0.5in}

\noindent
\textbf{(f)} [2 pts] What is the value of the F-statistic? Write your answer as 
a \emph{fraction}, not as a decimal. 

\vspace{0.5in}

<!-- \noindent -->
<!-- \textbf{(g)} [2 pts] The value of the F-statistic would be larger if the  -->
<!-- nitrate standard deviations were smaller for each group. Circle one.  -->

<!-- \begin{center} -->
<!-- \textbf{True} \hspace{2in} \textbf{False} -->
<!-- \end{center} -->

<!-- \vspace{0.25cm} -->

<!-- \noindent -->
<!-- \textbf{(h)} [2 pts] The value of the F-statistic would be larger if the  -->
<!-- nitrate means were more different across the groups. Circle one.  -->

<!-- \begin{center} -->
<!-- \textbf{True} \hspace{2in} \textbf{False} -->
<!-- \end{center} -->

<!-- \vspace{0.25cm} -->
\noindent
\textbf{(i)} [1 pt] Which distribution was used to obtain the p-value
presented in the table? Circle one. 

\begin{center}
\textbf{Simulated / Permuted Null Distribution} \hspace{1in}
\textbf{F-distribution}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(j)} [4 pts] Citing values from the ANOVA table to support your 
answer, what conclusions could be drawn regarding the hypotheses stated in (b)
and (c)? 

\vspace{1in}

\noindent
\textbf{(k)} [3 pts] The table below presents all comparisons of soil 
treatment. What value of $\alpha^*$ should the researchers use to determine
which of these tests produced "significant" results, so that the overall Type I
error rate for these tests is less than 5\%?

\vspace{0.1in}
\begin{center}
$\alpha^*= $
\end{center}
\vspace{0.1in}

```{r}
pairwise.t.test(lettuce$growth, lettuce$trt_group, 
                p.adjust.method = "none") %>% 
  tidy() %>% 
  mutate(group1 = fct_recode(group1, 
                             "Hydroponic B" = "hydro_B", 
                             "Soil" = "soil"
                             ), 
         group2 = fct_recode(group2, 
                             "Hydroponic A" = "hydro_A", 
                             "Hydroponic B" = "hydro_B"), 
         p.value = scales::pvalue(p.value, accuracy = 0.0001)
         ) %>% 
  rename(`Group 1` = group1, 
         `Group 2` = group2) %>% 
  pander()
```

\textbf{Extra Credit} [3 pts] Based on the comparisons above, sketch what you
would expect the side-by-side box-plots to look like. Be sure to label your
x-axis and y-axis!


\newpage

\noindent
\textbf{Q4} [19 points] As you may be aware, many individuals are concerned about
the presence of BPA in plastics, especially plastics that make contact with food
and drinks. Currently, these is an incomplete understanding of how exposure to
BPA affects our ingestion. Last year Dr. Hagobian in the Kinesiology and Public
Health carried out a study to investigate the role of Bisphenol A (BPA) in 
metabolism and endocrine disruption. 

Dr. Hahobian recruited 11 subjects, each of whom ate two types of cookies on two
separate visits, one visit in December and the second in February. On one visit
they ate the BPA-laced cookie and, on a different day, a placebo cookie (with no
BPA). Thirty minutes after eating the cookie on each occasion, they were given a
glucose tolerance test to measure their glucose metabolism. 

A summary of the glucose test results (mmol/L) after eating each type of cookie
as well as the difference in glucose results for each subject is shown below.

| Cookie                    | Mean   | Standard Deviation | n  |
|---------------------------|:------:|:-----------------:|:---:|
| Placebo                   |	5.259  | 0.762              | 11 |
| BPA	                      | 5.355	 | 1.462              |	11 |
| Difference: Placebo - BPA | -0.095 | 1.153              | 11 |

\vspace{0.5cm}

\noindent
\textbf{(a)} [2 pts] For simplicity, Dr. Hagobian could have given all
subjects the BPA cookie on their first visit in December, and the Placebo cookie
on the second visit in February, but he didn’t. Instead, when a subject came for
their first visit, he flipped a coin. If it was heads, they received BPA on that
visit (and Placebo on their second visit). If it was tails they received the 
Placebo cookie first. Why did he add this extra coin flipping step instead of
the simpler approach of just giving everyone one type of cookie in December and
the other type in February?

\vspace{1in}

\noindent
\textbf{(b)} [2 pts] Dr. Hagobian is interested in testing whether BPA causes a shift
in glucose levels. Which analysis would be more appropriate? Circle one. 

\vspace{0.25in}

\begin{center}
\textbf{Difference in Two Independent Means} \hspace{1in} 
\textbf{Mean of the Paired Differences}
\end{center}

\vspace{0.25in}

\noindent
\textbf{(c)} [3 pts] Based on your answer to (b), write out the null and 
alternative hypotheses for Dr. Hagobian's test using **notation**. *Be sure to 
indicate the order of subtraction being used!*

\vspace{0.25cm}

\textbf{$H_0$:}

\vspace{0.25in}

\textbf{$H_A$:}

\vspace{0.25in}

\newpage
\noindent
\textbf{(d)} [4 pts] To perform the analyses you selected in (b), what conditions does
Dr. Hagobian need to check before obtaining a p-value? Circle all that apply. 

\begin{multicols}{2}
\begin{itemize}
\item[(i)] Independence of the differences \\
\item[(ii)] Independence of the observations within each group \\
\item[(iii)] Independence of the observations between the groups \\
\item[(iv)] Independence of the variables \\
\item[(v)] Equal variance between the groups \\
\item[(vi)] Linear relationship between the variables \\
\item[(vii)] Normality of the differences \\
\item[(viii)] Normality of the observations within each group \\
\end{itemize}
\end{multicols}

\vspace{0.25cm}

\noindent
\textbf{(e)} [3 pts] Using `R`, Dr. Hagobian obtained the following table. 

```{r dr-hagobian-results}
dr_hagobian <- tibble(Difference = 
                        rnorm(n = 11, mean = -0.095, sd = 1.153)
                      )
dr_hagobian %>% 
  t_test(response = Difference, 
         alternative = "two-sided", 
         mu = 0, 
         conf_level = 0.90
         ) %>% 
  select(-t_df, -alternative) %>% 
  pander()

```

\vspace{0.25cm}

Which of the following would be the best overall conclusion in the context of 
Dr. Hagobian's study? Your selection should reflect the hypotheses you wrote in
part (c)!

\begin{itemize}
\item[(i)] With such a large p-value, we have significant evidence to reject the
null hypothesis. We conclude the true mean of the differences in glucose 
between eating a BPA cracker and a Placebo cracker is not 0. 

\item[(ii)] With such a large p-value, we have insufficient evidence to reject
the null hypothesis. We conclude the true mean of the differences in glucose 
between eating a BPA cracker and a Placebo cracker is 0. 

\item[(iii)] With such a large p-value, we have insufficient evidence to reject
the null hypothesis. We do not have evidence to suggest the mean of the
differences in glucose between eating a BPA cracker and a Placebo cracker is
different from 0. 

\item[(vi)] With such a large p-value, we have significant evidence to reject
the null hypothesis. We conclude the true mean glucose after eating a BPA
cracker is different from the true mean glucose after eating a Placebo cracker.

\item[(v)] With such a large p-value, we have insufficient evidence to reject
the null hypothesis. We conclude there is no difference in the true mean glucose
after eating a BPA cracker and the true mean glucose after eating a Placebo
cracker.

\item[(vi)] With such a large p-value, we have insufficient evidence to reject 
the null hypothesis. We do not have evidence to suggest the true mean glucose
after eating a BPA cracker is different from the true mean glucose after eating
a Placebo cracker.
\end{itemize}

\vspace{0.5in}

\noindent
\textbf{(f)} [2 pts] Based on the decision you reached in (e), what type of
error could you have made? Circle one. 

\begin{center}
\textbf{Type I Error} \hspace{1in} \textbf{Type II Error} \hspace{1in}
\textbf{No error was made} 
\end{center}

<!-- \vspace{0.25in} -->

<!-- \noindent -->
<!-- \textbf{(g)} [2 pts] If instead Dr. Hagobian had 100 subjects, the chance of  -->
<!-- the error described in part (f) would  -->

<!-- \begin{center} -->
<!-- \textbf{increase} \hspace{1in} \textbf{decrease} \hspace{1in} -->
<!-- \textbf{stay the same} -->
<!-- \end{center} -->

<!-- \newpage -->
\newpage

\noindent
\textbf{(h)} [2 pts] In a different study, Dr. Hagobian obtained a p-value of
0.0425 and a 95% confidence interval of (-1.129, 0.0437). Which of the 
following statements about these findings is true? Circle one.  

\begin{itemize}
\item[(i)] The results of the hypothesis test and the confidence interval tend to 
agree with each other at the 5\% significance level. Four percent of the time we
would obtain a statistic like the one we saw somewhere in the interval of 
-1.129 mmol/L to 0.0437 mmol/L. 

\item[(ii)] The results of the hypothesis test and the confidence interval are
conflicting at the 5\% significance level. With a p-value of 0.0425 we have 
evidence to reject the null hypothesis, which would mean that our confidence 
interval would not contain 0.

\item[(iii)] The results of the hypothesis test and the confidence interval are 
conflicting at the 5\% significance level. There's a 95\% chance that a
statistic of 0.0425 would be in the interval (-1.129, 0.0437). 

\item[(vi)] The results of the hypothesis test and the confidence interval seem to 
agree with one another at the 5\% significance level. With a p-value of 0.0425
we do not have evidence to reject the null hypothesis, thus indicating that 0
should be in our interval. 

\end{itemize}


\vspace{1in}

__Points Earned:__ \underline{\hspace{2in}}

__Total Points: 49__

