---
title: "STAT 218 -- Midterm 2"
date: "May 17, 2020"
author: "Dr. Allison Theobold"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(pander)
library(infer)
library(broom)
```

\begin{center}
\large{Name:} \underline{\hspace{3in}}
\vspace{0.25cm}

\large{Section (circle one):} \hspace{1cm} 8:10am \hspace{2cm} 9:10am 

\end{center}

\vspace{0.8cm}

\noindent {\bf Read and Sign the Following Statement:}

I understand that give or receiving help on this exam is a violation of academic
regulations and is punishable by a grade of __F__ in this course. This
includes looking at other students' exams and / or allowing other students,
actively or passively, to see answers on my exam. This also includes revealing,
actively or passively, any information about the exam to any member of 
Professor Theobold's STAT 218 class who has not yet taken the exam. The use of 
cell phones is strictly prohibited.

\begin{center}

\textbf{I pledge not to do any of these things.}

\vspace{0.5cm}

\textbf{Signed:} \underline{\hspace{4in}}

\end{center}

\vspace{1cm}

\noindent {\bf \textsf{Instructions.}}

\begin{itemize}
\item Read and sign the honesty pledge at the top of this page. Your exam will
not be graded unless the honesty pledge is signed!
\item You may use a calculator. You \textbf{may not} use your phone or any
device that connects to the internet as a calculator. \\
\item Show all work as clearly as possible. Point totals are shown in brackets 
next to each part. Formulas without values entered do not count as work. \\
\item All answers should be reported in decimal form, rounded to three decimal
places. \\
\item For multiple choice and multi-select problems, completely fill in the
provided circle (multiple choice) or square (multi-select) for your desired
answer choice(s). If you change an answer, be sure to completely erase your
initial selection.
\item You have 50 minutes to complete this exam, so budget your time wisely.    
\end{itemize}

\pagebreak

## Golden Ticket


| Scenario | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
|           |           |           |
| Summary measure | Mean | Slope or Correlation | Difference in Means |
|           |           |           |
| Parameter notation | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
|           |           |           |
| Statistic notation | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\bar{x}) = \frac{s}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\bar{x}_1 - \bar{x}_2) = \sqrt{\frac{s_1}{n_1} + \frac{s_2}{n_2}}$$

\newpage 

\textbf{Q2}[8 points] 
Is insomnia related to education level? Researchers at the University of 
Memphis, in Tennessee investigated this question in the *Journal of Abnormal
Psychology* (February, 2005). Eight hundred (800) adults living in Tennessee
were selected to participate in the study, using a random-digit telephone
dialing procedure. Responses were collected from a total of 575 adults. Two
variables measured were level of education and insomnia status (normal sleeper
or chronic insomniac). The researchers discovered that the lower the level of
education, the more likely the person was to have chronic insomnia.

\vspace{0.2cm}

\noindent
\textbf{(a)}[2 points] What are the observational units in this study?

\vspace{1in} 

\noindent
\textbf{(b)}[3 points] Fill in the blanks with one of the options provided in
parentheses to identify and explain the study design.

This is a(n) (observational study / randomized experiment)
\underline{\hspace{3in}} because (level of education / insomnia status) 
\underline{\hspace{3in}} (was / was not) \underline{\hspace{1in}} randomly
(assigned / sampled) \underline{\hspace{2in}}. 

\vspace{1cm}

\noindent
\textbf{(c)}[3 points] Which types of sampling bias may be present in this
study? Select **all** that might be present, or if there is likely to be no
bias present, select No bias.

\begin{itemize}
\item[(i)] Selection bias \\
\item[(ii)] Response bias \\
\item[(iii)] Non-response bias \\
\item[(iv)] No bias
\end{itemize}

\vspace{0.25in}

\textbf{Q5}[24 points] The *Journal of Food and Agriculture* contained an
article titled "Influence of hydroponic and soil cultivation on quality and
shelf life of ready-to-eat lamb's lettuce." In this article, researchers studied
the effects of different hydroponic growing methods on the nitrate content of
lettuce. In their study, the researchers randomly assigned 34 lettuce seedlings
to one of three growing conditions: soil, hydroponic A, or hydroponic B. At the
end of the growing period (60 days), nitrate measurements of the lettuce were
taken (mg / kg). 

```{r lettuce-sim-data}
lettuce <- tibble(
  trt_group = c(
    rep("soil", 9),
    rep("hydro_A", 12),
    rep("hydro_B", 13)
  ), 
  growth = c(
    rnorm(n = 9, mean = 3806, sd = 124), 
    rnorm(n = 12, mean = 4695, sd = 107), 
    rnorm(n = 13, mean = 3878, sd = 103)
    )
)
```

Results from the study are presented in the table below. 

```{r lettuce-table}
lettuce %>% 
  mutate(trt_group = fct_recode(trt_group, 
                                 "Hydroponic A" = "hydro_A", 
                                 "Hydroponic B" = "hydro_B", 
                                 Soil = "soil"
                                 ), 
         trt_group = fct_relevel(trt_group, 
                                 "Soil")
         ) %>% 
    group_by(trt_group) %>% 
  summarize(mean = mean(growth), 
            sd = sd(growth), 
            n = n()
            ) %>% 
  rename(`Treatment Group` = trt_group, 
         `Mean Growth` = mean, 
         `Standard Deviation of Growth` = sd, 
         `Sample Size` = n) %>% 
  pander()
```


\newpage

\noindent
\textbf{(a)}[2 points] One of the researcher's main questions was to determine
whether the growing method affects nitrate concentration in lettuce. Considering
how this study was executed, can they address this question? *Briefly justify
your answer.*

\vspace{1in}

Below is an incomplete ANOVA table, summarizing the data. You may use this 
information for the subsequent problems. 

```{r}
aov(growth ~ trt_group, data = lettuce) %>% 
  tidy() %>% 
  mutate(term = fct_recode(term, 
                           "Growing Method" = "trt_group"
                           ), 
         df = c("___", 31), 
         statistic = c("___", NA)
         ) %>% 
  pander()
  
```

\vspace{0.25in}

\noindent
\textbf{(b)}[3 points] In the context of the research question and in plain 
English, what are the null and alternative hypotheses being tested in the ANOVA
table above?

\vspace{0.25cm}
\noindent \textbf{$H_0$:}

\vspace{0.35in}

\noindent \textbf{$H_A$:}

\vspace{0.5in}

\noindent
\textbf{(c)}[2 points] Rewrite the null hypothesis above to use notation for the
\emph{parameters} that are being tested. 

\vspace{1in}

\noindent
\textbf{(d)}[2 points] The alternative hypothesis investigated in the ANOVA 
table above is:

$$H_A: \mu_{\text{Soil}} \neq \mu_{\text{Hydroponic A}} \neq 
\mu_{\text{Hydroponic B}}$$ 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(e)}[1 point] What are the degrees of freedom associated with 
`Growing Method`? 

\newpage

\noindent
\textbf{(f)}[2 points] What is the value of the F-statistic?  

\vspace{0.5in}

\noindent
\textbf{(g)}[2 points] The value of the F-statistic would be larger if the 
nitrate standard deviations were smaller for each group. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(h)}[2 points] The value of the F-statistic would be larger if the 
nitrate means were more different across the groups. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(i)}[1 point] Which distribution was used to obtain the p-value
presented in the table? Circle one. 

\begin{center}
\textbf{Simulated / Permuted Null Distribution} \hspace{1in}
\textbf{F-distribution}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(j)}[4 points] Citing values from the ANOVA table to support your 
answer, what conclusions could be drawn regarding the hypotheses stated in (b)? 

\vspace{1in}

\noindent
\textbf{(k)}[3 points] The table below presents all comparisons of the different
soil treatments. What value of $\alpha$ should the researchers use to determine
which, if any, of these tests produced "significant" results, so that the
overall Type I error rate for these tests is less than 5\%?

\vspace{0.5in}

```{r}
pairwise.t.test(lettuce$growth, lettuce$trt_group, 
                p.adjust.method = "none") %>% 
  tidy() %>% 
  mutate(group1 = fct_recode(group1, 
                             "Hydroponic B" = "hydro_B", 
                             "Soil" = "soil"
                             ), 
         group2 = fct_recode(group2, 
                             "Hydroponic A" = "hydro_A", 
                             "Hydroponic B" = "hydro_B"), 
         p.value = if_else(p.value < 5e-10, 
                           " < 0.000001", 
                           as.character(
                             round(p.value, 4)
                             )
                           )
         ) %>% 
  rename(`Group 1` = group1, 
         `Group 2` = group2) %>% 
  pander()
```

\textbf{Extra Credit}[3 points] Based on the comparisons above, sketch what you
would expect the side-by-side boxplots to look like. Be sure to label your
x-axis and y-axis!


\newpage

\textbf{Q3}[24 points] 
In 2004, the state of North Carolina released to the public a large dataset
containing information on births recorded in this state. This data set has been 
of interest to medical researchers who are studying the relation between habits
and practices of expectant mothers and the birth of their children. This 
analysis will focus on a random sample of 1,000 observations from the published
dataset. 

\vspace{0.25cm}

\noindent
\textbf{(a)}[3 points] Hospital administration at Duke University Hospital are
interested in the difference in the mean baby birth weight between mothers who
do not smoke and mothers do smoke. Using the table below, report the observed
statistic for this comparison. Indicate in your answer what notation should be
used for this statistic. 

\vspace{0.25cm}

\textbf{Answer:}

\vspace{0.5in}

\textbf{Notation:}

\vspace{0.25in}

```{r favstats-smoking}
favstats(weight ~ habit, data = births14) %>% 
  pander()
```

\vspace{0.5in}

```{r, out.width = "80%"}
births14 %>%
  filter(!is.na(habit)) %>% 
  ggplot(mapping = aes(x = weight)) + 
  geom_histogram(binwidth = 1.5) + 
  facet_wrap(~habit, scales = "free") + 
  theme_bw() +
  labs(x = "Weight of Baby (lbs)", 
       y = "Number of Babies")
```

\newpage

\noindent
\textbf{(b)}[4 points] These hospital administrators are interested in
\emph{estimating} the true difference in mean birth weight between
mothers who smoke and mothers who do not smoke. The administrators learned in
their Statistics class how obtain a confidence interval for a difference in
means using a $t$-distribution. Using the plots above evaluate whether it would
be appropriate for the administrators to use a $t$-distribution to obtain a
confidence interval for the true difference in means.

\vspace{1.5in}

\noindent
\textbf{(c)}[5 points] The administrators contacted the Department of Statistics
at Duke and requested a consultation. The Statistician they spoke with suggested
they use bootstrapping instead of a $t$-distribution to obtain a confidence
interval. Fill in the blanks below to explain to the administrators how one
bootstrap (re)sample is found. 

\begin{center}
\noindent
On \underline{\hspace{1in}} total cards, write \underline{\hspace{2in}} on the 
cards.
\end{center}

\vspace{0.25in}

Generate a new sample by: 

\vspace{0.5in}

\begin{center}
\noindent
Calculate and plot the \underline{\hspace{4cm}} from each bootstrap (re)sample. 
\end{center}

\vspace{0.25in}

\noindent
\textbf{(d)}[3 points] The Statistician carried out the above process to find
1000 bootstrap resamples using the order (`nonsmoker` - `smoker`) and obtained
the bootstrap distribution below. 

Where is the bootstrap distribution centered? Why is the distribution centered 
at that value?

```{r bootstrap-births, out.width = "40%", fig.align = 'left'}
bootstrap_dist <- births14 %>% 
  specify(response = weight, explanatory = habit) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", 
            order = c("nonsmoker", "smoker")
            )

bootstrap_dist %>% 
  visualise() + 
  labs(x = "Bootstrap Difference in Means (Nonsmoker - Smoker)", 
       title = "", 
       y = "Frequency") + 
  theme_bw()
```

\newpage

\noindent
\textbf{(e)}[3 points] The table below presents percentiles for the bootstrap 
distribution shown above. Circle the two values which will construct a 
\textbf{99\%} confidence interval. 

\vspace{0.25cm}

```{r percentile-CI-table}
bootstrap_dist %>% 
  summarize("0.5%" = quantile(stat, 0.005), 
            "1%" = quantile(stat, 0.01),
            "2.5%" = quantile(stat, 0.025),
            "5%" = quantile(stat, 0.05),
            "90%" = quantile(stat, 0.90),
            "95%" = quantile(stat, 0.95),
            "97.5%" = quantile(stat, 0.975),
            "99%" = quantile(stat, 0.99),
            "99.5%" = quantile(stat, 0.995)
            ) %>% 
  pivot_longer(cols = everything(), 
               names_to = "Quantile", 
               values_to = "Value") %>% 
  pander()
```

\vspace{0.25cm}

\noindent
\textbf{(f)}[4 points] Interpret the 99% confidence interval found in part (e)
\textbf{in the context of this study}. 

\vspace{1in}

\noindent
\textbf{(g)}[2 points] Based on your confidence interval in (e), which of the
following is the most likely p-value for a two-sided hypothesis test? Circle 
one. 

\begin{center}
(i) 0.20 \hspace{0.25in} (ii) 0.10 \hspace{0.25in} (iii) 0.05 \hspace{0.25in} 
(iv) 0.01 \hspace{0.25in} (v) 0.005
\end{center}
