---
title: 'Lab 9: Malaria'
author: "Present Group Member's Names Go Here!"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
## Load in the packages necessary for our analysis: tidyverse

library(tidyverse)
library(infer)
library(janitor)
```

An article that appeared in the journal Lancet in May of 2021 (Datoo, et al.)
described a study that investigated a potential vaccine that might protect 
children against malaria. Researchers recruited children between the ages of 5 
and 17 months in Burkina Faso, a country in western Africa, as participants. Th
e children were randomly assigned into one of three groups: one group received a
large dose of the potential vaccine, another group received a small dose, and a
third group received a placebo. Researchers observed the children for the next
18 months, keeping track of whether or not the child developed malaria.
Researchers hoped, of course, that children who received a vaccine would be less
likely to develop malaria than children who received a placebo.

## Setup

```{r, load-data}

malaria_data <- read_csv(here("data", "malaria_data.csv"))

malaria_data <- read_csv(here::here("two_days", 
                              "Week 9 - Two Cat Inference", 
                              "lab", 
                              "malaria_data.csv"))


```

**1. Click on the dataset to open the previewer. What is the observational unit for the study?**

**2. Which is the explanatory variable and which is the response variable? Specify the levels / units of each variable!**

-   Explanatory:

-   Response:

**3. Should you do a Chi-square test of Independence or test of Homogeneity for these data? Justify how you know.**

## Visualize & Summarize the Data

4.  Fill in the code below to make a **filled** bar plot of the data. What do you see in your filled-bar plot?

```{r filled-barplot}
ggplot(data = __________,
       mapping = aes(x = _________, fill = _______)) +
  geom_bar(position = "______") +
  labs(x = "____________",
       y = "____________")
```

**5.  Why do we want to make a filled bar plot instead of a stacked or dodged bar plot?**

6.  Finish the code below to make a two-way contingency table. Note: we want our **explanatory variable** to be indicated in the columns.

```{r two-way-table}
malaria_data |> 
   count(_____________, _____________) |> 
   pivot_wider(names_from = _______________, 
               values_from = n) |> 
   adorn_totals(where = c("row", "col"))

malaria_data |> 
   count(Malaria, Vaccine) |> 
   pivot_wider(names_from = Vaccine, 
               values_from = n) |> 
   adorn_totals(where = c("row", "col"))
```

**7. What proportion of children who received the high dose vaccine contracted malaria?**

**8. What proportion of children who received the low dose vaccine contracted malaria?**

**9. What proportion of children who received the placebo contracted malaria?**

## Theory-based Chi-square

**10. Write the null and alternative hypothesis for this study in words.**

-   Null:

-   Alternative:

Recall, in order for the $\chi^2$ distribution to be a good approximation of the true sampling distribution, we need to verify two conditions:

-   The observations are independent
-   We have a "large enough" sample size
    -   This is checked by verifying there are at least 5 expected counts in each cell

11. Is the independent observation condition violated? Justify your answer!


The equation for calculating expected counts is:

$$\frac{\text{row } i \text{ total count}\times \text{column } j \text{ total count}}{\text{total count}}$$ Our table is only a 2 x 3 table, but what if you
had a 6 x 8 or worse, 20 x 42 table? Checking each cell's expected count would
be very tedious. We only need to check the cell which will have the smallest
expected count.

12. Which row in your two-way contingency table from #6 has the smallest total count?

13. Which column in your two-way contingency table from #6 has the smallest total count?

**14. Using the equation above calculate the expected count for the cell in the row and column you specified in #9 and #10.**

**15. Is the "large enough" sample size condition violated?**

**16. Can we use the $\chi^2$ distribution to approximate the true sampling distribution?**

17. Fill in the code below to perform a theory-based Chi-square test of independence.

```{r chi-square-test-theory}
chisq_test(x = malaria_data, 
           response = ________________, 
           explanatory = ________________)
```

18. What conclusion would you reach based on your results? *Make sure to address (1) Chi-square test statistic and associated degrees of freedom, (2) p-value, (3)* $\alpha$ threshold, (4) your decision about the null hypothesis, and (5) your conclusion in context of the data.



## Simulated Chi-square

What if the large enough sample size condition was violated? We would have
needed to use a simulation-based approach instead of a theory-based approach! 
Let's walk through what this would look like.

**19. First, we need to calculate the observed chi-square test statistic from our data. We did this by hand during the activities, but it can be tedious so let's make R do it for us. Fill in the code below to calculate your observed chi-square test statistic.**

*Hint:* The statistic you want to calculate is abbreviated `"Chisq"`. 

```{r observed-chisq}
obs_xsq <- malaria_data |>
  specify(response = _____________________,
          explanatory = __________________) |> 
  calculate(stat = "_______")

```

20. What value of the observed chi-square test statistic did you obtain? Where have you previously seen this?

**21. Now we need to generate what the sampling distribution would look like if the null were true (aka null distribution of our chi-square statistics). Fill in the code below to generate and visualize the null distribution.**

```{r null-distribution}
null_dist <- malaria_data |>
  specify(response = Malaria, 
          explanatory = Vaccine) |> 
  hypothesize(null = "______________") |>
  generate(reps = 1000, type = "__________") |>
  calculate(stat = "Chisq")

visualize(data = null_dist, 
          method = "simulation")
```

**22. Once we have our null distribution, we can use this to calculate our simulated p-value.**

```{r simulated-pvalue}
get_pvalue(x = null_dist,
           obs_stat = obs_xsq,
           direction = "__________")
```

23. What conclusion would you reach with the simulated chi-square test? Does this differ from your answer in #17?

