## Activity 12:  Weather Patterns and Record Snowfall

\setstretch{1}

### Learning outcomes

* Given a research question involving one categorical explanatory variable and one quantitative response variable, construct the null and alternative hypotheses
  in words and using appropriate statistical symbols.

* Describe and perform a simulation-based hypothesis test for a difference in means.

* Interpret and evaluate a p-value for a simulation-based hypothesis test for a difference in means.

* Use bootstrapping to find a confidence interval for a difference in means.

* Interpret a confidence interval for a difference in means.

* Use a confidence interval to determine the conclusion of a hypothesis test.

### Terminology review

In today's activity, we will use simulation-based methods to analyze the association between one categorical explanatory variable and one quantitative response variable, where the groups formed by the categorical variable are independent. Some terms covered in this activity are:

* Independent groups

* Difference in means

To review these concepts, see Section 6.3 in the textbook.

### Weather patterns and record snowfall

In the winter of 2018--2019, Bozeman had a record snowfall which resulted in the collapse of two flat-roofed buildings on the MSU campus.  A writer for the *Washington Post* predicted the heavy snowfall for 2018--2019 due to the El Ni\latexcode{\~{n}}o weather pattern that occurred in that season. A meteorologist in Montana wanted to see if the weather pattern really was associated with total snowfall. She obtained historical data from 44 years on the weather pattern (El Ni\latexcode{\~{n}}o or La Ni\latexcode{\~{n}}a) and snowfall (in inches) at the Billings Weather Station [@weather].  Side-by-side boxplots and summary statistics for each group are shown on the following page.

Notice from the `R` code that the name of the data set is `Snow`.

```{r, echo=TRUE, collapse=FALSE}
# Read in data set
Snow <- read.csv("https://math.montana.edu/courses/s216/data/SnowfallByWeatherPattern.csv")
```

\newpage

```{r, echo = TRUE, out.width = "60%"}
# Side-by-side box plots
Snow %>%
ggplot(aes(x = WeatherPattern, y = Snowfall)) +
    geom_boxplot() + 
    labs(title = "Snowfall by weather pattern",
         x = "Weather pattern") +
    coord_flip()
```

```{r, echo=TRUE, collapse=FALSE}
# Summary statistics
Snow %>% 
     summarize(favstats(Snowfall ~ WeatherPattern))
```


#### Quantitative variables review  {-}

1. The two variables assessed in this study are the type of weather pattern and snowfall.  Identify the role for each variable (explanatory or response).

\vspace{.4in}

2. Which group (El Ni\latexcode{\~{n}}o or La Ni\latexcode{\~{n}}a) has the highest center in the distributions of snowfall? Explain which measure of center you are using.

\vspace{.4in}

3.  Using the side-by-side box plots, which group has the largest spread in snowfall?  How did you make that choice?

\vspace{.4in}

\newpage

4.  Is this an experiment or an observational study? Justify your answer.

\vspace{1in}

5.  Is this a paired data set or two independent groups?  Explain your reasoning.

\vspace{1in}

#### Ask a research question {-}

6.  Write out the parameter of interest in context of the study.  Use proper notation and be sure to define your subscripts.  Use El Ni\latexcode{\~{n}}o minus La Ni\latexcode{\~{n}}a as the order of subtraction.

\vspace{1in}

7.  Write out the null hypothesis in words.

\vspace{1in}

8.  Write the alternative hypothesis in notation. 

\vspace{0.5in}

#### Summarize and visualize the data {-}

9. Calculate the summary statistic of interest (difference in means).  Use El Ni\latexcode{\~{n}}o minus La Ni\latexcode{\~{n}}a as the order of subtraction. What is the appropriate notation for this statistic?

\vspace{0.5in}

\newpage 

#### Use statistical inferential methods to draw inferences from the data {-}

##### Hypothesis test {-}

Remember that the null distribution is created based on the assumption the null hypothesis is true.  In this study, the null hypothesis states that there is no association between the two variables.  This means that the snowfall values observed in the data set would have been the same regardless of the weather pattern that year.

To demonstrate this simulation, your instructor will provide cards to for you to use to simulate a sample.  

10.  How many cards will we start with?

\vspace{0.3in}

11.  What will we write on each card?

\vspace{0.3in}

12.  Next, we will mix the cards together and shuffle into two piles.  How many cards will go into each pile?  What should we label the piles?

\vspace{.8in}

13.  What value is calculated from the cards and plotted on the null distribution?  *Hint*: What statistic are we calculating from the data?

\vspace{0.3in}

14.  Create one simulation using the cards provided.  Is your simulated statistic closer to the null value of zero than the difference in means calculated from the sample?  Explain why this makes sense.

\vspace{0.8in}

15.  Once we create a null distribution of 1000 simulations, at what value do you expect the distribution to be centered?  Explain your reasoning.

\vspace{.8in}

We will use the `two_mean_test()` function in `R` (in the `catstats` package) to simulate the null distribution of differences in sample means and compute a p-value. 

\newpage 

16.  When using the `two_mean_test()` function, we need to enter the name of the response variable, `Snowfall`, and the name of the explanatory variable, `WeatherPattern`, for the formula.  The name of the data set as shown above is `Snow`.  What values should be entered for each of the following to create 1000 simulated samples?

* First in subtraction (What is the outcome for the explanatory variable that is used as first in the order of subtraction? `"El_Nino"` or `"La_Nina"`):

\vspace{.2in}
* Number of repetitions:
    
\vspace{.2in}
* As extreme as:
    
\vspace{.2in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.2in}

17.  Simulate a null distribution and compute the p-value. Using the `R` script file for this activity, enter your answers for question 16 in place of the `xx`'s to produce the null distribution with 1000 simulations.  Highlight and run lines 1--29.

```{r, echo = TRUE, eval = FALSE}
two_mean_test(Snowfall ~ WeatherPattern, data = Snow,  # Variables and data
         first_in_subtraction = "xx", # First outcome in order of subtraction
         number_repetitions = 1000,  # Number of simulations
         as_extreme_as = xx,  # Observed statistic
         direction = "xx")  # Direction of alternative: "greater", "less", or "two-sided"
```

|        Sketch the null distribution created using the code above.
\vspace{1.5in}


18.  Report the p-value. Based off of this p-value, write a conclusion to the hypothesis test.

\vspace{0.9in}

\newpage

##### Confidence interval {-}

We will use the `two_mean_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of differences in sample means and calculate a confidence interval. 

19. Using bootstrapping find a 95\% confidence interval. Using the provided `R` script file, enter the variable names and data set name as in the `two_mean_test()` function, outcome name for the first in subtraction, number of repetitions, and the confidence level as a decimal.  Highlight and run lines 32--35. Report the 95\% confidence interval in interval notation.
```{r, echo = TRUE, eval = FALSE}
two_mean_bootstrap_CI(RESPONSE ~ EXPLANATORY, data = DATASET,  # Variables and data
                      first_in_subtraction = "xx", # First value in order of subtraction
                      number_repetitions = 1000,  # Number of simulations
                      confidence_level = xx)
```

\vspace{0.3in}

20. Interpret the interval you calculated in question 19. 

\vspace{1in}


21.  Would the results from a theory-based test match the results we saw with the simulation?  Explain why or why not.

\vspace{1in}

\newpage

### Take-home messages

1.	This activity differs from Activities 11a and 11b because the responses are independent, not paired.  These data are analyzed as a difference in means, not a mean difference.  

2. To create one simulated sample on the null distribution for a difference in sample means, label cards with the response variable values from the original data.  Mix cards together and shuffle into two new groups of sizes $n_1$ and $n_2$. Calculate and plot the difference in means. 

3. To create one simulated sample on the bootstrap distribution for a difference in sample means, label $n_1 + n_2$ cards with the original response values.  Keep groups separate and randomly draw with replacement $n_1$  times from group 1 and $n_2$ times from group 2.  Calculate and plot the resampled difference in means.  


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered

\newpage

