---
title: "Final Exam -- Group Portion"
output: html_document
---

```{r setup, include = FALSE}
# This is code to make your final report look nice! 
# Please don't modify these options! 
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")

library(tidyverse)
library(infer)
library(janitor)
```

## Investigating the Relationship between Helmet Use and Head Injuries

Although using a helmet is assumed to reduce the risk of head injuries in alpine
sports, this effect is questioned. In contrast to bicycling or inline skating,
there is no policy of mandatory helmet use for recreational alpine skiers and
snowboarders. Sullheim et al. (2017) investigated the relationship between 
helmet use and head injury, at eight major Norwegian ski resorts during the 
2022 winter season.

An injury was recorded when a skier or snowboarder was treated by the ski patrol
or first aid room. A total of 725 skier or snowboarder injuries were recorded.
As a control group, 2992 noninjured skiers or snowboarders were interviewed at
the same eight resorts. Every 10th skier or snowboarder waiting in line was
interviewed to achieve a systematic sample of the population. The interviews
were done every Wednesday and Saturday during the 4 winter months of 2002 by
personnel who were not told the purpose of the study.

```{r data, include = FALSE}
helmet <- read_csv(here::here(
  "Week 10 - Review & Final", 
  "final", 
  "group", 
  "data",
  "HeadInjuries.csv")
  )

helmet <- read_csv(here::here(
  "data",
  "HeadInjuries.csv")
  )

```

# Exploratory Data Analysis

## Visualizing the Frequency of Injuries

__1. Let's start by creating a bar plot, visualizing the frequency of head injuries in the dataset.__

```{r injury-bar-plot}

```

__2. Describe what you see in the bar plot.__



## Visualizing the Frequency of Helmet Use

__3. Next, let's visualize the frequency of helmet use in the dataset.__ 

```{r helmet-bar-plot}

```

__4. Describe what you see in the bar plot.__



## Visualizing the Relationship between Helmet Use and Head Injury

__5. Create a filled bar plot visualizing the relationship between helmet use and head injury.__

```{r helmet-injury-barplot}

```

__6. Describe what you see in the bar plot.__



## Summarizing with Statistics

Now that we've visualized the data, let's calculate some summary statistics.

__7. Create a table summarizing the frequency of people who were / were not injured.__

```{r injury-table}

```

__8. Create a table summarizing the frequency of people who did / did not wear a helmet.__

```{r helmet-table}

```

I've written the code to produce the two-way table for you. Run the code to
answer the subsequent questions. 

```{r two-way-table}
helmet %>% 
  group_by(Helmet, Outcome) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  pivot_wider(names_from = Outcome, 
              values_from = n) %>% 
  adorn_totals(where = c("row", "col")) 
```

__9. What proportion of people wearing a helmet had head injuries?__



__10. What proportion of people not wearing a helmet had head injuries?__


## Performing a Chi-Square Analysis

__11. Assess if you believe the condition of independence is violated. Justify your answer!__


__12. Assess if you believe the condition of a large enough sample size is violated. Justify your answer!__


__13. Fill in the code below to carry out a Chi-Squared analysis.__

```{r chi-sq}
chisq_test()
```

__14. Based on the analysis above, write a conclusion for the hypothesis test.__



## Assessing the Risk of Head Injury

__15. Calculate the relative risk of a head injury, comparing the non-helmet wearers to the helmet wearers.__ 



__16. Interpret this value in the context of this study.__

