---
title: "Midterm 1 -- Group Portion"
output: html_document
---

```{r setup, include = FALSE}
# This is code to make your report format look nice! 
# Please don't modify these options! 
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")

library(tidyverse)
library(mosaic)
library(moderndive)
```

## Investigating hours worked

For this exam, we'll be working with data from Lab 3 --- the General Social
Survey (GSS). The GSS is a high-quality survey which gathers data on American 
society and opinions, conducted since 1972. This data set is a sample of 500
entries from the GSS, spanning years 1973-2018, including demographic markers
and some economic variables.

```{r data-load}
GSS <- read_csv(here::here("Week 5 - Difference in Means & Midterm 1", 
                           "midterm 1",
                           "group", 
                           "data",
                           "GSS_clean.csv"
                           )
                )


GSS <- read_csv(here::here("data", 
                           "GSS_clean.csv"
                           )
                )
```

## Investigating Education & Spousal Education

You've explored regression before, both in Lab 4 and in your Midterm Project. 
For this exam, we will investigate if / how much the highest year of school a 
person's spouse completed (`highest_year_school_completed_spouse`) can
predict a person's education (`highest_year_of_school_completed`). 

## Summarizing the Variables

__1. Using the `favstats()` function, find summary statistics for the highest year of school each person in this sample completed.__

```{r education-summary}
favstats(____, data = GSS)
```

__2. Using the `favstats()` function, find summary statistics for the highest year of school each person's spouse completed.__

```{r spouse-summary}
favstats(____, data = GSS)
```


## Visualizing the Relationship

Next, let's visualize the relationship between the highest year of school a 
person's spouse completed (`highest_year_school_completed_spouse`) and the
highest year of school their spouse completed 
(`highest_year_of_school_completed`). To be clear, 
`highest_year_school_completed_spouse` is our **explanatory** variable and 
`highest_year_of_school_completed` is our **response** variable. 

__3. I've started the code for you, but you will need to add:__

- the variables
- the points
- nice axis labels

```{r scatterplot}
ggplot(data = GSS, 
       mapping = aes(x = ____, y = ____)) 
```

__4. Describe the relationship you see in the plot. Be sure to address the shape, direction and strength of the relationship.__


__5. Are there any outliers present in these data? If so, where?__ 


__6. If we don't remove the outliers you mentioned, what will happen to the regression line?__


## Linear Regression

Alright, the relationship between `highest_year_of_school_completed` and 
`highest_year_school_completed_spouse` appears reasonably linear, so let's
model this relationship with a line. 


