---
title: "Midterm 1 Question Bank"
author: "Stat 218"
output: 
  pdf_document:
    extra_dependencies: ["multicol", "setspace"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(moderndive)
library(pander)
library(infer)
library(lterdatasampler)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted, so it is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours.

\vspace{0.5cm}

## Provided Formulas

$$R^2 = r^2 = \frac{s_{y}^2 - s_{residuals}^2}{s_y^2}$$ 

\vspace{0.25cm}

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$

\newpage

\noindent {\textbf Q1} Dr. John Arbuthnot, an 18th century physician, writer, 
and mathematician is famous for performing the first hypothesis test of
significance. Dr. Arbuthnot was interested in the ratio of newborn boys to
newborn girls, so he gathered the baptism records for children born in London
for every year from 1629 to 1710. Artbuthnot found that in every year, the
number of males born in London exceeded the number of females.

\noindent
\textbf{(a)} [2 pts] Describe the sampling method used by Dr. Arbuthnot.

\vspace{0.5in}

\noindent 
\textbf{(b)} [2 pts] Describe how this sampling method could be biased.

\vspace{0.5in}

\noindent 
\textbf{(c)} [3 pts] A preview of the dataset is provided below. Use this
preview to address the following questions. 

```{r data-preview}
arbuthnot
```

\begin{itemize}
\item Identify the cases in the data set. \\

\vspace{1cm}

\item List the variables. Indicate whether each variable is categorical or
quantitative. If the variable is quantitative, give the units. \\

\vspace{1cm}

\item What would the dimensions of the data set be? (number of rows by number of
columns)

\end{itemize}

\vspace{2cm}

\newpage

\textbf{(d)} [3 pts] A scatterplot displaying the number of girls born over time
is displayed below. Describe the relationship you see in the scatterplot. Be
sure to address the form, direction, strength, and outliers present in the 
scatterplot.

```{r arbuthnot-scatter}
ggplot(data = arbuthnot, 
       aes(x = year, y = girls)) + 
  geom_point() + 
  labs(x = "Year", 
       y = "Number of Girls")
```

\vspace{2cm}

\textbf{(d)} [3 pts] Would it be appropriate to model the relationship between
the number of girls born and the year with a linear regression? Justify your
belief! 

\newpage

\noindent {\textbf Q2} I collected data on 512 different fast food items from
Mcdonalds, Chick-Fil-A, Sonic, Arby's, Burger King, Dairy Queen, Subway, and
Taco Bell. For each restaurant, I sampled 64 items from their menu and recorded
the nutritional content of each item (e.g., calories, saturated fat, calcium, 
protein, etc.). 

\noindent
\textbf{(a)} [2 pts] Describe the sampling method I used to obtain these 512 
fastfood items. 

\vspace{2cm}

\noindent
\textbf{(b)} [3 pts] I am interested in studying the relationship between the
total calories of a food item and the amount of saturated fat that item
contains.

__Write the null hypothesis for my question of interest.__

\vspace{2cm}

\noindent
\textbf{(c)} [2 pts] Is the alternative hypothesis one- or two-sided? Select
one.

\begin{itemize} 
\item One-sided \\
\item Two-sided
\end{itemize}

\noindent
\textbf{(d)} [5 pts] On the following page is the plot of the simulated null
distribution from `R`. Fill in the blanks below with one answer in each set of
parentheses to correctly explain how one sample on the null distribution would
be created.  Blanks preceded by (#) should be filled in with a number. All other
blanks should be filled in with the context of the study.

\begin{center}
On (\#) \underline{\hspace{2in}} cards, write \underline{\hspace{2in}} on the 
cards.

\vspace{0.5cm}

Assume the null hypothesis is true and \underline{\hspace{3in}}.

\vspace{0.5cm}

Generate a new sample of 512 ordered pairs by \underline{\hspace{3in}}.

\vspace{0.5cm}

Calculate and plot the (slope/correlation) \underline{\hspace{4cm}} from each 
simulated sample.

\end{center}

```{r null-dist-fastfood}

obs_slope <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  calculate(stat = "slope") %>% 
  pull()
  
null_dist <- fastfood %>% 
  specify(response = sat_fat, 
          explanatory = calories) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, 
           type = "permute") %>% 
  calculate(stat = "slope")

visualise(null_dist) +
  labs(x = "Simulated Slope Statistic", 
       y = "Frequency (out of 1000 reps)", 
       title = "") +
  xlim(c(-0.015, 0.015))
```

\noindent
\textbf{(e)} [2 pts] Using the regression output below, draw a vertical line
where the observed statistic falls on the null distribution. 

```{r}
fastfood_lm <- lm(sat_fat ~ calories, data = fastfood)

get_regression_table(fastfood_lm) %>% 
  pander()
```

\vspace{0.25cm}

\noindent
\textbf{(f)} [2 pts] Shade the location of the plot you would use to calculate
the p-value. 

\vspace{0.25cm}

\noindent
\textbf{(g)} [1 pts] Estimate the p-value associated with this hypothesis test. 

\vspace{0.5cm}

\noindent
\textbf{(h)} [3 pts] Based on your estimated p-value, what would you conclude 
for this hypothesis test? (circle one)

\begin{itemize}
\item In less than 1 out of 1000 simulated samples, we would observe a sample
slope of 0.017 or more extreme, if there is no linear relationship between 
the total calories and the saturated fat of a fast food item. 

\item If there is a linear relationship between the total calories and the
saturated fat of a fast food, we would observe a sample slope of 0.017 or more
extreme with a probability of less than 1 out of 1000.

\item The probability of seeing a sample slope between the total calories and
the saturated fat of a fast food item of 0.017 or more extreme is less than 0.1%. 

\item  The probability that there is no linear relationship between the total
calories and the saturated fat of a fast food item, is less than 0.1%.
\end{itemize}

\newpage

\textbf{Q3} The Atlantic marsh fiddler crab, Minuca pugnax (formerly Uca
pugnax), lives in salt marshes throughout the eastern coast of the United 
States. Historically, M. pugnax were distributed from northern Florida to Cape
Cod, Massachusetts, but like other species have expanded their range northward
due to ocean warming.

The Plum Island Ecosystem Long Term Ecological Research site collected data on
fiddler crabs from 13 marshes on the Atlantic coast of the United States in
summer 2016. The marshes spanned from northeast Florida to northeast
Massachusetts. Researchers were able to collect between 25 and 37 adult male
fiddler crabs at each marsh. 

\noindent 
\textbf{(a)} [3 pts] A preview of the dataset is provided below. Use this
preview to address the following questions. 

```{r crab-preview}
pie_crab %>% 
  select(-name, -air_temp_sd, -water_temp_sd) %>% 
  glimpse()
```

\begin{itemize}
\item Identify the cases in the data set. \\

\vspace{1cm}

\item List the variables. Indicate whether each variable is categorical or
quantitative. If the variable is quantitative, give the units. \\

\vspace{2cm}

\item What would the dimensions of the data set be? (number of rows by number of
columns)
\end{itemize}

\vspace{0.5cm}

\textbf{(b)} [3 pts] A histogram displaying the size of the sample of fiddler
crabs is displayed below. Describe the shape of the distribution. Be sure to 
address the center, spread, shape, and outliers. 

```{r}
pie_crab %>% 
  ggplot(aes(x = size)) +
  geom_histogram(binwidth = 1.5, color = "white") +
  labs(x = "Carapace Width of Fiddler Crab (millimeters)", 
       y = "Number of Crabs")
```

\vspace{1in}

\textbf{(c)} [2 pts] When using a $t$-distribution to find a 95% confidence 
interval, how many degrees of freedom should be used?

\vspace{1cm}

```{r}
crab_ci <- t_test(pie_crab, 
       response = size,
       conf_int = TRUE, 
       conf_level = 0.95) %>% 
  select(estimate, lower_ci, upper_ci)
```

\textbf{(d)} [4 pts] A 95% confidence interval for the mean carapace width of a
Fiddler Crab was found to be (`r round(crab_ci$lower_ci, 2)`, 
`r round(crab_ci$upper_ci, 2)`). Below is the researchers' interpretation of
this confidence interval: 

\begin{center}
\emph{There is a 95\% chance that that every sample of Fiddler Crabs will have
a mean carapace width between 14.31 and 15.01 millimeters.}
\end{center}

\noindent
Identify __two__ mistakes committed and fix them. Be brief but clear in your
description.

\vspace{.15in}

Mistake 1: \underline{\hspace{6in}}

\vspace{0.5cm}

Fix: \underline{\hspace{6in}}

\vspace{0.5cm}

Mistake 2: \underline{\hspace{6in}}

\vspace{0.5cm}

Fix: \underline{\hspace{6in}}

\newpage

\textbf{(f)} [2 points] Can the researchers use their interval to make
inferences about __all__ fiddler Crabs in the United States? Justify your 
answer!

\vspace{1in}

\noindent \textbf{Q4} Researchers are interested in the fish that reside in the
Caspian Sea. They have plans to collect many fish and take multiple measurements
on each. Match each statistical description on the right with each piece of
information given. Put the letter of the statistical description in the blanks
on the left.

\begin{center}
	\begin{tabular}{p{8cm}p{6cm}}
		\underline{\hspace{.4in}} circumference of the fish &	(a) quantitative variable \\
		&   \\
		\underline{\hspace{.4in}} species of the fish																																															&	(b) categorical variable \\
		&   \\
		\underline{\hspace{.4in}} average length of all fish in \\ the area of consideration																													&		(c) parameter: $\mu$	\\
		&   \\
		\underline{\hspace{.5in}} mean internal temperature \\ of the fish collected in the sample																										&	(d) statistic: $\bar{x}$	\\
		&   \\
		\underline{\hspace{.4in}} one of the fish in the area of \\ consideration																																		&	(e) observational unit	\\
		&   \\
		\underline{\hspace{.4in}} method of only studying the fish \\ caught in the net \@ 3pm on Wednesday of the research time frame								&	(f) cluster sampling method	\\
		&   \\
		\underline{\hspace{.4in}} method of selecting 5\% of each \\ species, known to be in the area of consideration, for the sample								&	(g) stratified sampling method	\\
		&   \\
		\underline{\hspace{.3in}} method of dividing up the whole \\ location with netting and sampling 10 random netted areas												& (h) convenience sampling method
	\end{tabular}
\end{center}

\vspace{0.5cm}

\noindent \textbf{Q5} [4 points] Indicate whether each statement about a bootstrap sample 
is true or false.

\begin{itemize}
\item The resample and original sample MUST be the same size. \\

\item The resample and original sample BOTH are taken from the population. \\

\item The resample can ONLY use values that were in the original sample. \\

\item The resample uses ALL values that were in the original sample. \\
\end{itemize}

\newpage

\noindent \textbf{Q6} The purpose of creating a bootstrap distribution is to:

\begin{enumerate}
\item Estimate the mean of the unknown sampling distribution.
\item Estimate the spread of the unknown sampling distribution.
\item Both A and B.
\item None of the above.
\end{enumerate}

\vspace{0.25cm}

\noindent \textbf{Q7} [2 points] A newspaper article claims that the average age
for people who receive food stamps in a community is 40 years. A local
researcher believes that the average age is less than that. The researcher takes
a random sample of 100 people in the community who receive food stamps, and
finds their average age to be 39.2 years, which is statistically significantly
lower than the age of 40 stated in the article (p-value < .05). Indicate for
each of the following interpretations whether they are valid or invalid.

\noindent \textbf{(a)} The statistically significant result indicates that the 
majority of people who receive food stamps is younger than 40.
\begin{enumerate}
\item[(a)] Valid
\item[(b)] Invalid
\end{enumerate}

\vspace{0.25cm}

\noindent \textbf{(b)} An error must have been made. This difference in means
(39.2 vs. 40 years) is too small to be statistically significant.
\begin{enumerate}
\item[(a)] Valid
\item[(b)] Invalid
\end{enumerate}

\newpage

\noindent \textbf{Q6} Normal human body temperature is said to be 98.6 degrees
Fahrenheit.  A researcher studying body temperatures took a random sample of 50 
people and recorded the body temperatures for each.  The average body
temperature of the 50 people was 98.26.  Using these data, estimate the true
average body temperature of people. 

\textbf{(a)} Define the parameter of interest, using proper notation. 

\vspace{1in}

\textbf{(b)} What type of inference should be done to answer the research
question? (circle one)

\begin{itemize}
\item Confidence Interval
\item Hypothesis Test
\item No inference is needed. The true average body temperature is 98.6.
\item No inference is needed.  The true average body temperature is 98.26.
\end{itemize}

\textbf{(c)} The bootstrap distribution with 1001 bootstrap resamples is
below. What does one dot on the plot represent in the context of this problem?


\textbf{(d)} Explain how you could use cards to create __one__ bootstrap
resample.

\textbf{(e)} The percentile method is used in the picture to give a confidence
interval of (98.081, 98.512).  What level of confidence is used for this
interval?

\textbf{(f)} How many bootstrap resamples would be cut-off in the left-tail of
the distribution to create a 95% confidence interval? (circle one)

\begin{itemize}
\item 50
\item 25
\item 5
\item 10
\end{itemize}

\textbf{(g)} Use the plot and t* = 1.677 to create a 90% confidence interval
using the t*SE method.  Show your work.

\textbf{(h)} Interpret your 90% confidence interval from e.

\textbf{(i)} Based off your interval in (h), can we be 90% sure that the true
average body temperature is lower than 98.6 degrees Fahrenheit?  Explain.

