
5.  Write out the null hypothesis in proper notation for this study.  

\vspace{0.8in}

6.  What sign ($<$, $>$, or $\neq$) would you use in the alternative hypothesis for this study?  Explain your choice.

\vspace{0.5in}

Upload and open the `R` script file for Week 11 lab. Upload and import the csv file, `swearing`. Enter the name of the data set (see the environment tab) for datasetname in the `R` script file in line 6. Write a title for the boxplots in line 14. Highlight and run lines 1--15 to load the data and create plots of the data. You may use the arrows on the plots tab to view both graphical representations of the data.

```{r, echo=TRUE, eval=FALSE}
swearing <- datasetname
paired_observed_plot(swearing)

swearing_diff <- swearing %>% 
  mutate(differences = Swear - Neutral)
swearing_diff %>%
  ggplot(aes(x = differences))+
  geom_boxplot()+
  labs(title="title",
       x = "Differences in pain tolerance (swearing condition - neutral condition)")
```

7. Based on the plots, does there appear to be some evidence in favor of the alternative hypothesis?  How do you know?
\vspace{0.4in}

Highlight and run lines 17--18 to get the summary statistics for the data.

```{r, echo=TRUE, eval=FALSE}
swearing_diff %>% 
    summarise(favstats(differences))
```

8.  What is the value of $\bar{x}_d$?  What is the sample size?
\vspace{0.25in}

9. **How far, on average, is each difference in pain tolerance from the mean of the differences in pain tolerance?  What is the appropriate notation for this value?**

\vspace{0.5in}

### Use statistical inferential methods to draw inferences from the data {-}

10.  Using the provided graphs and summary statistics, determine if both theory-based methods and simulation methods could be used to analyze the data.  Explain your reasoning.

\vspace{0.8in}

### Hypothesis test {-}

Remember that the null distribution is created based on the assumption the null hypothesis is true.  In this study, the null hypothesis states that swearing does not affect pain tolerance, or that the length of time a subject kept their hand in the water would be the same whether the patient was swearing or not.

We will use the `paired_test()` function in `R` (in the `catstats` package) to simulate the null distribution of sample means differences and compute a p-value. 

11.  When using the `paired_test()` function, we need to enter the name of the data set, either the order of subtraction (if the data set has both measurements) or the name of the differences (if the data set contains them).  We will also need to provide `R` with the observed mean difference, the direction of the alternative hypothesis, and the shift required in order to force the null hypothesis to be true.  The name of the data set as shown above is `swearing_diff` and the column of differences is called `differences`.  What values should be entered for each of the following to create 1000 simulated samples?

\vspace{1mm}
* shift:

\vspace{.2in}
* As extreme as:
    
\vspace{.2in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.2in}
* Number of repetitions:
    
\vspace{.2in}

12.  Simulate a null distribution and compute the p-value. Using the `R` script file for this lab, enter your answers for question 11 in place of the `xx`'s to produce the null distribution with 1000 simulations.  Highlight and run lines 21--27.  

```{r, echo = TRUE, eval = FALSE}
paired_test(data = swearing$differences,   # Vector of differences 
                                 # or data set with column for each group
        shift = xx,   # Shift needed for bootstrap hypothesis test
        as_extreme_as = xx,  # Observed statistic
        direction = "xx",  # Direction of alternative
        number_repetitions = xx,  # Number of simulated samples for null distribution
        which_first = 1)  # Not needed when using calculated differences
```
|        Sketch the null distribution created using the `paired_test` code.

\vspace{1.5in}

### Communicate the results and answer the research question {-}

13.  **Report the p-value. Based off of this p-value and a 1% significance level, what decision would you make about the null hypothesis?  What potential error might you be making based on that decision?**

\vspace{0.5in}

14. Do you expect the 98\% confidence interval to contain the null value of zero?  Explain.

\vspace{0.8in}

### Confidence interval {-}

We will use the `paired_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of sample mean differences and calculate a confidence interval. 

15. Using bootstrapping and the provided `R` script file, find a 98\% confidence interval. Fill in the missing values/numbers in the `paired_bootstrap_CI()` function to create the 98\% confidence interval.  Highlight and run lines 30--33. **Upload a copy of the bootstrap distribution created to Gradescope for your group.** 
```{r, echo = TRUE, eval = FALSE}
paired_bootstrap_CI(data = swearing_diff$differences, # Enter vector of differences
                    number_repetitions = 1000, # Number of bootstrap samples for CI
                    confidence_level = xx,  # Confidence level in decimal form
                    which_first = 1)  # Not needed when entering vector of differences
```
\newpage

|        Sketch the bootstrap distribution created using the code.  Report the 98\% confidence interval in interval notation.

\vspace{1.5in}

16. Interpret the *confidence level* of the interval you calculated in question 15. 

\vspace{0.8in}

17.  Write a paragraph summarizing the results of this study as if you were describing the results to your roommate.  **Upload a copy of your group's paragraph to Gradescope.** Be sure to describe:

* Summary statistic

* P-value and interpretation

* Conclusion (written to answer the research question)

* Confidence interval and interpretation

* Scope of inference

\vspace{2.6in}

\newpage