---
title: "Midterm 2 Question Bank"
author: "Stat 218"
output:
  pdf_document:
    extra_dependencies:
    - multicol
    - setspace
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(moderndive)
library(pander)
library(infer)
library(lterdatasampler)
library(palmerpenguins)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted. It is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours prior to the exam.

\vspace{0.5cm}

## Golden Ticket


| Scenario | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
| Summary measure | Mean | Slope or Correlation | Difference in Means |
| Parameter notation | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
| Statistic notation | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\mu_1 - \mu_2) = \sqrt{\frac{\sigma_1}{n_1} + \frac{\sigma_2}{n_2}}$$

\newpage 

\textbf{Q1}[10 points] 

Researchers in Southern England collected data on grassland butterflies. They
were interested in whether movement patterns varied across species and between
male and female butterflies. Researchers observed 164 butterflies over the
three-year length of this study, of which 28 were female and 136 were male. 
These 164 butterflies are considered to be representative of all grassland
butterflies. The butterfly movements were observed by measuring how far they
flew (in meters) from one landing site to the next, called step distance. This
was done by placing a flag at each landing site and measuring the distance
between the flags using a mapping software.

The researchers were interested in investigating if there was a difference
between how far male and female butterflies travel, on average, between landing
sites. 

\noindent
\textbf{(a)}[4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.2cm}

Step distance is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}. 

\vspace{0.2cm} 

Sex is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}.

\vspace{0.25cm}

\noindent
\textbf{(b)}[3 pts] Which type of plot is most appropriate to display the
relationship between step distance and sex of the butterfly? Select one. 

\begin{itemize}
\item Segmented Bar plot \\
\item Scatterplot \\
\item Side-by-side Boxplot \\
\item Histogram 
\end{itemize}

\noindent
\textbf{(c)} [3 pts] Assuming a statistical difference in step distance is
found between the males and females in the sample, what is the scope of
inference for this study? Select one. 

\begin{itemize}
\item Sex causes a difference in average step distance for all grassland
butterflies. \\
\item Sex is associated with a difference in average step distance for grassland
butterflies similar to those sampled. \\
\item Sex causes a difference in average step distance for grassland butterflies
similar to those sampled. \\
\item Sex is associated with a difference in average step distance for all
grassland butterflies. 
\end{itemize}

\newpage 

\noindent
\textbf{Q2}[8 points] 
Is insomnia related to education level? Researchers at the University of 
Memphis, in Tennessee investigated this question in the Journal of Abnormal
Psychology (February, 2005). Eight hundred (800) adults living in Tennessee were
selected to participate in the study, using a random-digit telephone dialing
procedure. Responses were collected from a total of 575 adults. Two variables
measured were level of education and insomnia status (normal sleeper or chronic
insomniac). The researchers discovered that the lower the level of education,
the more likely the person was to have chronic insomnia.

\vspace{0.2cm}

\noindent
\textbf{(a)}[2 points] What are the observational units in this study?

\vspace{1cm} 

\noindent
\textbf{(b)}[3 points] Fill in the blanks with one of the options provided in
parentheses to identify and explain the study design.

This is a(n) (observational study / randomized experiment)
\underline{\hspace{3in}} because (level of education / insomnia status) \underline{\hspace{3in}} (was / was not) \underline{\hspace{3in}} randomly
(assigned / sampled) \underline{\hspace{3in}. 

\vspace{1cm}

\noindent
\textbf{(c)}[3 points] Which types of sampling bias may be present in this
study? Select **all** that are present, or if there is no bias present, select
"No bias." 

\begin{itemize}
\item Selection bias \\
\item Response bias \\
\item Non-response bias \\
\item No bias
\end{itemize}

\newpage

\noindent
\textbf{Q3}[23 points] 
In 2004, the state of North Carolina released to the public a large dataset
containing information on births recorded in this state. This data set has been 
of interest to medical researchers who are studying the relation between habits
and practices of expectant mothers and the birth of their children. This 
analysis will focus on a random sample of 1,000 observations from the published
dataset. 

\noindent
\textbf{(a)}[3 points] Researchers are interested in the difference in the mean
baby birth weight between mothers who smoke and mothers who do not smoke. Using
the table below, report the observed statistic for this comparison. Indicate in
your answer what notation should be used for this statistic. 

\vspace{0.25cm}

\textbf{Answer:}

\vspace{0.5in}

\textbf{Notation:}

\vspace{0.25cm}

```{r favstats-smoking}
favstats(weight ~ habit, data = births14)
```

\newpage 

```{r}
births14 %>%
  filter(!is.na(habit)) %>% 
  ggplot(mapping = aes(x = weight)) + 
  geom_histogram(binwidth = 1.5) + 
  facet_wrap(~habit, scales = "free") + 
  theme_bw() +
  labs(x = "Weight of Baby (lbs)", 
       y = "Number of Babies")
```

\noindent
\textbf{(b)}[4 points] Hospital administration at Duke University Hospital are
interested in estimating the true difference in mean birth weight between
mothers who smoke and mothers who do not smoke. The administrators learned in
their Statistics class how obtain a confidence interval for a difference in
means using a $t$-distribution. Using the plots above and your knowledge of how
the data were collected evaluate whether it would be appropriate to use a
$t$-distribution to obtain a confidence interval. 

\vspace{1.5in}

\noindent
\textbf{(c)}[5 points] The administrators contacted the Department of Statistics
at Duke and requested a consultation. The Statistician they spoke with suggested
they use bootstrapping instead of a $t$-distribution to obtain a confidence
interval. Fill in the blanks below to explain to the adminstrators how one
bootstrap (re)sample is found. 

\begin{center}
On (\#) \underline{\hspace{2in}} cards, write \underline{\hspace{2in}} on the 
cards.

\vspace{0.5cm}

Generate a new sample by: \underline{\hspace{3in}}.

\vspace{0.5cm}

Calculate and plot the \underline{\hspace{4cm}} from each bootstrap (re)sample. 

\end{center}

\noindent
\textbf{(d)}[3 points] The Statistician carried out a simulation with 1000 
bootstrap resamples using the order (`nonsmoker` - `smoker`) and obtained the
bootstrap distribution below. 

Where is the bootstrap distribution centered? Why is the distribution centered 
at that value?

\vspace{1in}

```{r bootstrap-births}
bootstrap_dist <- births14 %>% 
  specify(response = weight, explanatory = habit) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", 
            order = c("nonsmoker", "smoker")
            )

bootstrap_dist %>% 
  visualise() + 
  labs(x = "Bootstrap Difference in Means (Nonsmoker - Smoker)", 
       title = "", 
       y = "Frequency")
```


\noindent
\textbf{(e)}[3 points] Use the table below to find the 99% confidence interval
using the percentile method. 

```{r percentile-CI-table}
bootstrap_dist %>% 
  summarize("0.5%" = quantile(stat, 0.05), 
            "1%" = quantile(stat, 0.01),
            "2.5%" = quantile(stat, 0.025),
            "5%" = quantile(stat, 0.05),
            "90%" = quantile(stat, 0.90),
            "95%" = quantile(stat, 0.95),
            "97.5%" = quantile(stat, 0.975),
            "99.5%" = quantile(stat, 0.95)
            ) %>% 
  pivot_longer(cols = everything(), 
               names_to = "Quantile", 
               values_to = "Value") %>% 
  pander()
```

\textbf{99\% Interval:}

\vspace{0.25cm}

\noindent
\textbf{(f)}[5 points] Interpret the 99% confidence interval found in part (f). 

\vspace{1in}

\noindent
\textbf{Q4}[] As you may be aware, many individuals are concerned about the
presence of BPA in plastics, especially plastics that make contact with food and
drinks. Currently, these is an incomplete understandinf of how exposure to BPA
affects our ingestion. Last year Dr. Hagobian in the Kinesiology and Public
Health carried out a study to investigate the role of Bisphenol A (BPA) in 
metabolism and endocrine disruption. 

Dr. Hahobian recruited 11 subjects, each of whom ate two types of cookies on two
separate visits, one visit in December and the second in February. On one visit
they ate the BPA-laced cookie and, on a different day, a placebo cookie (with no
BPA). Thirty minutes after eating the cookie on each occasion, they were given a glucose tolerance test to measure their glucose metabolism as well as other
blood tests to measure levels of estrogen, pro-insulin, and c-peptide in the
blood. 

A summary of the glucose test results (mmol/L) after eating each type of cookie
as well as the difference in glucose results for each subject is shown below.

| Cookie                    | Mean   | Standard Deviation | n  |
|---------------------------|:------:|:-----------------:|:---:|
| Placebo                   |	5.259  | 0.762              | 11 |
| BPA	                      | 5.355	 | 1.462              |	11 |
| Difference: Placebo - BPA | -0.095 | 1.153              | 11 |


\noindent
\textbf{(a)}[4 points] For simplicity, Dr. Hagobian could have given all
subjects the BPA cookie on their first visit in December, and the Placebo cookie
on the second visit in February, but he didn’t. Instead, when a subject came for
their first visit, he flipped a coin. If it was heads, they received BPA on that
visit (and Placebo on their second visit). If it was tails they received the 
Placebo cookie first. Why did he add this extra coin flipping step instead of
the simpler approach of just giving everyone one type of cookie in December and
the other in February?

\vspace{1in}

\noindent
\textbf{(b)}[2 points] Dr. Hagobian is interested in testing whether BPA causes a shift
in glucose levels. Which analysis would be more appropriate? Circle one. 

\textbf{Difference in Two Independent Means} \hspace{1in} \textbf{Mean of the Differences}


\noindent
\textbf{(c)}[2 points] Based on your answer to (b), write out the null and 
alternative hypotheses for Dr. Hagobian's test. 

\vspace{0.25cm}

\textbf{$H_0$:}

\vspace{0.25in}

\textbf{$H_A$:}

\vspace{0.25in}


\noindent
\textbf{(d)}[] To perform the analyses you selected in (b), what conditions does
Dr. Hagobian need to check before obtaining a p-value? Select all that apply. 

\begin{itemize}
\item Independence of the differences \\
\item Independence of observations within each group \\
\item Independence of observations between the groups \\
\item Independence of the variables \\
\item Equal variance between the groups \\
\item Linear relationship between the variables \\
\item Normality of the Differences \\
\item Normality of the observations within each group \\
\end{itemize}

\vspace{0.25cm}

\noindent
\textbf{(e)}[] Using `R`, Dr. Hagobian obtained the following table. 

```{r}
dr_hagobian <- tibble(Difference = 
                        rnorm(n = 11, mean = -0.095, sd = 1.153)
                      )
dr_hagobian %>% 
  t_test(response = Difference, 
         alternative = "two-sided", 
         mu = 0, 
         conf_level = 0.90
         ) %>% 
  select(-t_df, -alternative) %>% 
  pander()

```

\vspace{0.25cm}

Which of the following would be the best overall conclusion in the context of 
Dr. Hagobian's study? Your selection should reflect the hypotheses you wrote in
part (c)!

\begin{itemize}
\item With such a large p-value, we have significant evidence to reject the
null hypothesis that the difference in the glucose levels after eating a BPA
and Placebo cracker are the same. We can therefore conclude that eating a BPA
cracker shifted the glucose levels.  

\item
\end{itemize}


<!-- Question about p-value calculation.  -->

\noindent
\textbf{(f)}[] Based on the decision you reached in (e), what type of error 
could you have made? Circle one. 

\begin{itemize}
\item Type I error \\
\item Type II error \\
\item No error was made \\
\end{itemize}


\noindent
\textbf{(g)}[] If instead Dr. Hagobian had 100 subjects (50 in each group), 
would the chance of the error described in part (f) *increase*, *decrease*, or
*stay the same*. 



