---
title: "Midterm 2 Question Bank"
author: "Stat 218"
output:
  pdf_document:
    extra_dependencies: ["multicol", "setspace"]
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")

set.seed(1234)

library(tidyverse)
library(openintro)
library(mosaic)
library(pander)
library(infer)
library(forcats)
library(broom)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted. It is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours prior to the exam.

\vspace{0.5cm}

## Golden Ticket


| Scenario | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
| Summary measure | Mean | Slope or Correlation | Difference in Means |
| Parameter notation | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
| Statistic notation | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\mu_1 - \mu_2) = \sqrt{\frac{\sigma_1}{n_1} + \frac{\sigma_2}{n_2}}$$

\newpage 

\textbf{Q1}[10 points] Researchers in Southern England collected data on
grassland butterflies. They
were interested in whether movement patterns varied across species and between
male and female butterflies. Researchers observed 164 butterflies over the
three-year length of this study, of which 28 were female and 136 were male. 
These 164 butterflies are considered to be representative of all grassland
butterflies. The butterfly movements were observed by measuring how far they
flew (in meters) from one landing site to the next, called step distance. This
was done by placing a flag at each landing site and measuring the distance
between the flags using a mapping software.

The researchers were interested in investigating if there was a difference
between how far male and female butterflies travel, on average, between landing
sites. 

\noindent
\textbf{(a)}[4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.2cm}

Step distance is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}. 

\vspace{0.2cm} 

Sex is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}.

\vspace{0.25cm}

\noindent
\textbf{(b)}[3 pts] Which types of plot is most appropriate to display the
relationship between step distance and sex of the butterfly? Select all that 
apply. 

\begin{itemize}
\item[(i)] Segmented Bar plot \\
\item[(ii)] Scatterplot \\
\item[(iii)] Side-by-side Boxplot \\
\item[(iv)] Faceted histograms 
\end{itemize}

\noindent
\textbf{(c)} [3 pts] Assuming a statistical difference in step distance is
found between the males and females in the sample, what is the scope of
inference for this study? Select one. 

\begin{itemize}
\item[(i)] Sex causes a difference in average step distance for all grassland
butterflies. \\
\item[(ii)] Sex is associated with a difference in average step distance for grassland
butterflies similar to those sampled. \\
\item[(iii)] Sex causes a difference in average step distance for grassland butterflies
similar to those sampled. \\
\item[(iv)] Sex is associated with a difference in average step distance for all
grassland butterflies. 
\end{itemize}

\newpage 

\noindent
\textbf{Q2}[8 points] 
Is insomnia related to education level? Researchers at the University of 
Memphis, in Tennessee investigated this question in the Journal of Abnormal
Psychology (February, 2005). Eight hundred (800) adults living in Tennessee were
selected to participate in the study, using a random-digit telephone dialing
procedure. Responses were collected from a total of 575 adults. Two variables
measured were level of education and insomnia status (normal sleeper or chronic
insomniac). The researchers discovered that the lower the level of education,
the more likely the person was to have chronic insomnia.

\vspace{0.2cm}

\noindent
\textbf{(a)}[2 points] What are the observational units in this study?

\vspace{1cm} 

\noindent
\textbf{(b)}[3 points] Fill in the blanks with one of the options provided in
parentheses to identify and explain the study design.

This is a(n) (observational study / randomized experiment)
\underline{\hspace{3in}} because (level of education / insomnia status) 
\underline{\hspace{3in}} (was / was not) \underline{\hspace{3in}} randomly
(assigned / sampled) \underline{\hspace{3in}. 

\vspace{1cm}

\noindent
\textbf{(c)}[3 points] Which types of sampling bias may be present in this
study? Select **all** that are present, or if there is no bias present, select
"No bias." 

\begin{itemize}
\item[(i)] Selection bias \\
\item[(ii)] Response bias \\
\item[(iii)] Non-response bias \\
\item[(iv)] No bias
\end{itemize}

\newpage

\noindent
\textbf{Q3}[24 points] 
In 2004, the state of North Carolina released to the public a large dataset
containing information on births recorded in this state. This data set has been 
of interest to medical researchers who are studying the relation between habits
and practices of expectant mothers and the birth of their children. This 
analysis will focus on a random sample of 1,000 observations from the published
dataset. 

\vspace{0.25cm}

\noindent
\textbf{(a)}[3 points] Researchers are interested in the difference in the mean
baby birth weight between mothers who smoke and mothers who do not smoke. Using
the table below, report the observed statistic for this comparison. Indicate in
your answer what notation should be used for this statistic. 

\vspace{0.25cm}

\textbf{Answer:}

\vspace{0.5in}

\textbf{Notation:}

\vspace{0.25in}

```{r favstats-smoking}
favstats(weight ~ habit, data = births14)
```

\vspace{0.5in}

```{r, out.width = "80%"}
births14 %>%
  filter(!is.na(habit)) %>% 
  ggplot(mapping = aes(x = weight)) + 
  geom_histogram(binwidth = 1.5) + 
  facet_wrap(~habit, scales = "free") + 
  theme_bw() +
  labs(x = "Weight of Baby (lbs)", 
       y = "Number of Babies")
```

\newpage

\noindent
\textbf{(b)}[4 points] Hospital administration at Duke University Hospital are
interested in estimating the true difference in mean birth weight between
mothers who smoke and mothers who do not smoke. The administrators learned in
their Statistics class how obtain a confidence interval for a difference in
means using a $t$-distribution. Using the plots above and your knowledge of how
the data were collected evaluate whether it would be appropriate to use a
$t$-distribution to obtain a confidence interval for true difference in means.

\vspace{1.5in}

\noindent
\textbf{(c)}[5 points] The administrators contacted the Department of Statistics
at Duke and requested a consultation. The Statistician they spoke with suggested
they use bootstrapping instead of a $t$-distribution to obtain a confidence
interval. Fill in the blanks below to explain to the administrators how one
bootstrap (re)sample is found. 

\begin{center}
\noindent
On (\#) \underline{\hspace{1in}} cards, write \underline{\hspace{2in}} on the 
cards.
\end{center}

\vspace{0.25in}

Generate a new sample by: 

\vspace{0.5in}

\begin{center}
\noindent
Calculate and plot the \underline{\hspace{4cm}} from each bootstrap (re)sample. 
\end{center}

\vspace{0.25in}

\noindent
\textbf{(d)}[3 points] The Statistician carried out a simulation with 1000 
bootstrap resamples using the order (`nonsmoker` - `smoker`) and obtained the
bootstrap distribution below. 

Where is the bootstrap distribution centered? Why is the distribution centered 
at that value?

```{r bootstrap-births, out.width = "60%", fig.align = 'left'}
bootstrap_dist <- births14 %>% 
  specify(response = weight, explanatory = habit) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", 
            order = c("nonsmoker", "smoker")
            )

bootstrap_dist %>% 
  visualise() + 
  labs(x = "Bootstrap Difference in Means (Nonsmoker - Smoker)", 
       title = "", 
       y = "Frequency") + 
  theme_bw()
```

\newpage

\noindent
\textbf{(e)}[3 points] The table below presents percentiles for the bootstrap 
distribution above. Circle the two values which will construct a 99% confidence
interval. 

\vspace{0.25cm}

```{r percentile-CI-table}
bootstrap_dist %>% 
  summarize("0.5%" = quantile(stat, 0.05), 
            "1%" = quantile(stat, 0.01),
            "2.5%" = quantile(stat, 0.025),
            "5%" = quantile(stat, 0.05),
            "90%" = quantile(stat, 0.90),
            "95%" = quantile(stat, 0.95),
            "97.5%" = quantile(stat, 0.975),
            "99.5%" = quantile(stat, 0.95)
            ) %>% 
  pivot_longer(cols = everything(), 
               names_to = "Quantile", 
               values_to = "Value") %>% 
  pander()
```

\vspace{0.25cm}

\noindent
\textbf{(f)}[4 points] Interpret the 99% confidence interval found in part (f)
\textbf{in the context of this study}. 

\vspace{1in}

\noindent
\textbf{(g)}[2 points] Based on your confidence interval in (f), which of the
following is the most likely p-value for a two-sided hypothesis test? Circle 
one. 

\begin{center}
(i) 0.20 \hspace{0.25in} (ii) 0.10 \hspace{0.25in} (iii) 0.05 \hspace{0.25in} 
(iv) 0.01 \hspace{0.25in} (v) 0.005
\end{center}

\vspace{0.25in}

\noindent
\textbf{Q4}[21 points] As you may be aware, many individuals are concerned about
the presence of BPA in plastics, especially plastics that make contact with food
and drinks. Currently, these is an incomplete understanding of how exposure to
BPA affects our ingestion. Last year Dr. Hagobian in the Kinesiology and Public
Health carried out a study to investigate the role of Bisphenol A (BPA) in 
metabolism and endocrine disruption. 

Dr. Hahobian recruited 11 subjects, each of whom ate two types of cookies on two
separate visits, one visit in December and the second in February. On one visit
they ate the BPA-laced cookie and, on a different day, a placebo cookie (with no
BPA). Thirty minutes after eating the cookie on each occasion, they were given a
glucose tolerance test to measure their glucose metabolism as well as other
blood tests to measure levels of estrogen, pro-insulin, and c-peptide in the
blood. 

A summary of the glucose test results (mmol/L) after eating each type of cookie
as well as the difference in glucose results for each subject is shown below.

| Cookie                    | Mean   | Standard Deviation | n  |
|---------------------------|:------:|:-----------------:|:---:|
| Placebo                   |	5.259  | 0.762              | 11 |
| BPA	                      | 5.355	 | 1.462              |	11 |
| Difference: Placebo - BPA | -0.095 | 1.153              | 11 |

\newpage

\noindent
\textbf{(a)}[4 points] For simplicity, Dr. Hagobian could have given all
subjects the BPA cookie on their first visit in December, and the Placebo cookie
on the second visit in February, but he didn’t. Instead, when a subject came for
their first visit, he flipped a coin. If it was heads, they received BPA on that
visit (and Placebo on their second visit). If it was tails they received the 
Placebo cookie first. Why did he add this extra coin flipping step instead of
the simpler approach of just giving everyone one type of cookie in December and
the other in February?

\vspace{1in}

\noindent
\textbf{(b)}[2 points] Dr. Hagobian is interested in testing whether BPA causes a shift
in glucose levels. Which analysis would be more appropriate? Circle one. 

\begin{center}
\textbf{Difference in Two Independent Means} \hspace{1in} 
\textbf{Mean of the Paired Differences}
\end{center}

\vspace{0.25in}

\noindent
\textbf{(c)}[3 points] Based on your answer to (b), write out the null and 
alternative hypotheses for Dr. Hagobian's test using **notation**. *Be sure to 
indicate the order of subtraction being used!*

\vspace{0.25cm}

\textbf{$H_0$:}

\vspace{0.25in}

\textbf{$H_A$:}

\vspace{0.25in}

\noindent
\textbf{(d)}[3 points] To perform the analyses you selected in (b), what conditions does
Dr. Hagobian need to check before obtaining a p-value? Circle all that apply. 

\begin{multicols}{2}
\begin{itemize}
\item[(i)] Independence of the differences \\
\item[(ii)] Independence of observations within each group \\
\item[(iii)] Independence of observations between the groups \\
\item[(iv)] Independence of the variables \\
\item[(v)] Equal variance between the groups \\
\item[(vi)] Linear relationship between the variables \\
\item[(vii)] Normality of the differences \\
\item[(viii)] Normality of the observations within each group \\
\end{itemize}
\end{multicols}

\newpage

\noindent
\textbf{(e)}[3 points] Using `R`, Dr. Hagobian obtained the following table. 

```{r dr-hagobian-results}
dr_hagobian <- tibble(Difference = 
                        rnorm(n = 11, mean = -0.095, sd = 1.153)
                      )
dr_hagobian %>% 
  t_test(response = Difference, 
         alternative = "two-sided", 
         mu = 0, 
         conf_level = 0.90
         ) %>% 
  select(-t_df, -alternative) %>% 
  pander()

```

\vspace{0.25cm}

Which of the following would be the best overall conclusion in the context of 
Dr. Hagobian's study? Your selection should reflect the hypotheses you wrote in
part (c)!

\begin{itemize}
\item[(i)] With such a large p-value, we have significant evidence to reject the
null hypothesis. We conclude the true mean of the differences in glucose 
between eating a BPA cracker and a Placebo cracker is not 0. 

\item[(ii)] With such a large p-value, we have insufficient evidence to reject the
null hypothesis. We conclude the true mean of the differences in glucose 
between eating a BPA cracker and a Placebo cracker is 0. 

\item[(iii)] With such a large p-value, we have insufficient evidence to reject the
null hypothesis. We do not have evidence to suggest the mean of the differences
in glucose between eating a BPA cracker and a Placebo cracker is different 
from 0. 

\item[(vi)] With such a large p-value, we have significant evidence to reject the
null hypothesis. We conclude the true mean glucose after eating a BPA cracker 
is different from the true mean glucose after eating a Placebo cracker.

\item[(v)] With such a large p-value, we have insufficient evidence to reject the
null hypothesis. We conclude there is no difference in the true mean glucose
after eating a BPA cracker and the true mean glucose after eating a Placebo
cracker.

\item[(vi)] With such a large p-value, we have insufficient evidence to reject the
null hypothesis. We do not have evidence to suggest the true mean glucose after
eating a BPA cracker is different from the true mean glucose after eating a
Placebo cracker.
\end{itemize}

\vspace{0.25in}

\noindent
\textbf{(f)}[2 points] Based on the decision you reached in (e), what type of
error could you have made? Circle one. 

\begin{center}
\textbf{Type I Error} \hspace{1in} \textbf{Type II Error} \hspace{1in}
\textbf{No error was made} 
\end{center}

\vspace{0.25in}

\noindent
\textbf{(g)}[2 points] If instead Dr. Hagobian had 100 subjects (50 in each
group), the chance of the error described in part (f) would 

\begin{center}
\textbf{increase} \hspace{1in} \textbf{decrease} \hspace{1in}
\textbf{stay the same}
\end{center}

\newpage

\noindent
\textbf{(h)}[2 points] In a different study, Dr. Hagobian obtained a p-value of
0.0425 and a 95% confidence interval of (-1.129, 0.0437). Which of the 
following statements about these findings is true? Circle one.  

\begin{itemize}
\item[(i)] The results of the hypothesis test and the confidence interval tend to 
agree with each other at the 5% significance level. Four percent of the time we
would obtain a statistic like the one we saw somewhere in the interval of 
-1.129 mmol/L to 0.0437 mmol/L. 

\item[(ii)] The results of the hypothesis test and the confidence interval are
conflicting at the 5% significance level. With a p-value of 0.0425 we have 
evidence to reject the null hypothesis, which would mean that our confidence 
interval would not contain 0.

\item[(iii)] The results of the hypothesis test and the confidence interval are 
conflicting at the 5% significance level. There's a 95% chance that a statistic
of 0.0425 would be in the interval (-1.129, 0.0437). 

\item[(vi)] The results of the hypothesis test and the confidence interval seem to 
agree with one another at the 5% significance level. With a p-value of 0.0425
we do not have evidence to reject the null hypothesis, thus indicating that 0
should be in our interval. 

\end{itemize}

\noindent
\textbf{Q5}[24 points] The *Journal of Food and Agriculture* contained an
article titled "Influence of hydroponic ad soil cultivation on quality and shelf
life of ready-to-eat lamb's lettuce." In this article, researchers studied the
effects of different hydroponic growing methods on the nitrate content of
lettuce. In their study, the researchers randomly assigned 34 lettuce seedlings
to one of three growing conditions: soil, hydroponic A, or hydroponic B. At the
end of the growing period (60 days), nitrate measurements of the lettuce were
taken (mg / kg). 

```{r lettuce-sim-data}
lettuce <- tibble(
  trt_group = c(
    rep("soil", 9),
    rep("hydro_A", 12),
    rep("hydro_B", 13)
  ), 
  growth = c(
    rnorm(n = 9, mean = 3806, sd = 124), 
    rnorm(n = 12, mean = 4695, sd = 107), 
    rnorm(n = 13, mean = 3878, sd = 103)
    )
)
```

Results from the study are presented in the table below. 

```{r lettuce-table}
lettuce %>% 
  mutate(trt_group = fct_recode(trt_group, 
                                 "Hydroponic A" = "hydro_A", 
                                 "Hydroponic B" = "hydro_B", 
                                 Soil = "soil"
                                 ), 
         trt_group = fct_relevel(trt_group, 
                                 "Soil")
         ) %>% 
    group_by(trt_group) %>% 
  summarize(mean = mean(growth), 
            sd = sd(growth)) %>% 
  rename(`Treatment Group` = trt_group, 
         `Mean Growth` = mean, 
         `Standard Deviation of Growth` = sd) %>% 
  pander()
```


\noindent
\textbf{(a)}[2 points] One of the researcher's main questions was to determine
whether the growing method affects nitrate concentration in lettuce. Considering
how this study was executed, can they address this question? *Briefly justify
your answer.*

\vspace{1in}

Below is an incomplete ANOVA table, summarizing the data. You may use this 
information for the subsequent problems. 

```{r}
aov(growth ~ trt_group, data = lettuce) %>% 
  tidy() %>% 
  mutate(term = fct_recode(term, 
                           "Growing Method" = "trt_group"
                           ), 
         df = c("___", 31), 
         statistic = c("___", NA)
         ) %>% 
  pander()
  
```

\newpage

\noindent
\textbf{(b)}[3 points] In the context of the research and in plain English, 
what are the null and alternative hypotheses investigated in the ANOVA table 
above?

\vspace{0.25cm}
\noindent \textbf{$H_0$:}

\vspace{0.35in}

\noindent \textbf{$H_A$:}

\vspace{0.5in}

\noindent
\textbf{(c)}[2 points] Rewrite the null hypothesis above to use notation for the
parameters that are being tested. 

\vspace{1in}

\noindent
\textbf{(d)}[2 points] The alternative hypothesis investigated in the ANOVA 
table above is $H_A$: $\mu_{\text{Soil}} \neq \mu_{\text{Hydroponic A}} \neq 
\mu_{\text{Hydroponic B}}$. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(e)}[1 point] What are the degrees of freedom associated with 
`Growing Method`? 

\vspace{0.5in}

\noindent
\textbf{(f)}[2 points] What is the value of the F-statistic?  

\vspace{0.5in}

\noindent
\textbf{(g)}[2 points] The value of the F-statistic would be larger if the 
nitrate standard deviations were smaller for each group. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(h)}[2 points] The value of the F-statistic would be larger if the 
nitrate means were more different across the groups. Circle one. 

\begin{center}
\textbf{True} \hspace{2in} \textbf{False}
\end{center}

\vspace{0.25cm}
\noindent
\textbf{(i)}[1 point] Which distribution was used to obtain the p-value
presented in the table? Circle one. 

\begin{center}
\textbf{Simulated / Permuted Null Distribution} \hspace{1in}
\textbf{F-distribution}
\end{center}

\newpage

\noindent
\textbf{(j)}[4 points] Citing values from the ANOVA table to support your 
answer, what conclusions could be drawn regarding the hypotheses stated in (b)
and (c)? 

\vspace{1in}

\noindent
\textbf{(k)}[3 points] The table below presents all comparisons of soil 
treatment. What value of $\alpha$ should the researchers use to determine
which of these tests produced "significant" results?

\vspace{0.5in}

```{r}
pairwise.t.test(lettuce$growth, lettuce$trt_group, 
                p.adjust.method = "none") %>% 
  tidy() %>% 
  mutate(group1 = fct_recode(group1, 
                             "Hydroponic B" = "hydro_B", 
                             "Soil" = "soil"
                             ), 
         group2 = fct_recode(group2, 
                             "Hydroponic A" = "hydro_A", 
                             "Hydroponic B" = "hydro_B"), 
         p.value = if_else(p.value < 5e-10, 
                           " < 0.000001", 
                           as.character(
                             round(p.value, 4)
                             )
                           )
         ) %>% 
  rename(`Group 1` = group1, 
         `Group 2` = group2) %>% 
  pander()
```

