---
title: "Midterm 2 Question Bank"
author: "Stat 218"
output:
  pdf_document:
    extra_dependencies:
    - multicol
    - setspace
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center", 
                      fig.width = 4, 
                      fig.height = 4)

library(tidyverse)
library(openintro)
library(mosaic)
library(moderndive)
library(pander)
library(infer)
library(lterdatasampler)
library(palmerpenguins)
```

Your exam questions will be randomly selected from this bank of
questions. There \textbf{will not} be a solution key posted. It is your 
responsibility to discuss your ideas with your group members and / or with 
Dr. Theobold during office hours prior to the exam.

\vspace{0.5cm}

## Golden Ticket


| Scenario | One Quantitative Response | Two Quantitative Variables | Quant. Response and Categ. Explanatory | 
|----------|:-------------------------:|:--------------------------:|:---------------------------------------:|
| Type of plot | Dot plot, Histogram, Boxplot | Scatterplot | Faceted Histograms, Side-by-side Boxplots |
| Summary measure | Mean | Slope or Correlation | Difference in Means |
| Parameter notation | $\mu$ | Slope: $\beta_1$; Correlation: $\rho$ | $\mu_1 - \mu_2$ |
| Statistic notation | $\bar{x}$ | Slope: $b_1$; Correlation: $r$ | $\bar{x}_1 - \bar{x}_2$ | 

\vspace{0.5cm}

## Provided Formulas

$$IQR = Q3 - Q1$$

\vspace{0.25cm}

__1.5 IQR Rule:__ above $Q3 + (1.5 \times IQR)$  or below $Q1 - (1.5 \times IQR)$

\vspace{0.25cm}

$$\hat{y} = b_0 + b_1 \times x$$

\vspace{0.2cm}

$$Residual = y - \hat{y}$$ 

\vspace{0.3cm}

__t-based confidence interval:__ \hspace{1cm} $\text{point estimate} \pm t^*_{df} \times SE$

\vspace{0.25cm}

$$SE(\mu) = \frac{\sigma}{\sqrt{n}}$$
\vspace{0.25cm}

$$SE(\mu_1 - \mu_2) = \sqrt{\frac{\sigma_1}{n_1} + \frac{\sigma_2}{n_2}}$$

\newpage 

\textbf{Q1}[10 points] 

Researchers in Southern England collected data on grassland butterflies. They
were interested in whether movement patterns varied across species and between
male and female butterflies. Researchers observed 164 butterflies over the
three-year length of this study, of which 28 were female and 136 were male. 
These 164 butterflies are considered to be representative of all grassland
butterflies. The butterfly movements were observed by measuring how far they
flew (in meters) from one landing site to the next, called step distance. This
was done by placing a flag at each landing site and measuring the distance
between the flags using a mapping software.

The researchers were interested in investigating if there was a difference
between how far male and female butterflies travel, on average, between landing
sites. 

\noindent
\textbf{(a)}[4 pts] Fill in each blank with one of the options in parentheses to
best describe the variables collected. 

\vspace{0.2cm}

Step distance is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}. 

\vspace{0.2cm} 

Sex is the (explanatory/response) \underline{\hspace{3in}} and it is
(categorical/quantitative) \underline{\hspace{3in}}.

\vspace{0.25cm}

\noindent
\textbf{(b)}[3 pts] Which type of plot is most appropriate to display the
relationship between step distance and sex of the butterfly? Select one. 

\begin{itemize}
\item Segmented Bar plot \\
\item Scatterplot \\
\item Side-by-side Boxplot \\
\item Histogram 
\end{itemize}

\noindent
\textbf{(c)} [3 pts] Assuming a statistical difference in step distance is
found between the males and females in the sample, what is the scope of
inference for this study? Select one. 

\begin{itemize}
\item Sex causes a difference in average step distance for all grassland
butterflies. \\
\item Sex is associated with a difference in average step distance for grassland
butterflies similar to those sampled. \\
\item Sex causes a difference in average step distance for grassland butterflies
similar to those sampled. \\
\item Sex is associated with a difference in average step distance for all
grassland butterflies. 
\end{itemize}

\newpage 

\noindent
\textbf{Q2}[8 points] 
Is insomnia related to education level? Researchers at the University of 
Memphis, in Tennessee investigated this question in the Journal of Abnormal
Psychology (February, 2005). Eight hundred (800) adults living in Tennessee were
selected to participate in the study, using a random-digit telephone dialing
procedure. Responses were collected from a total of 575 adults. Two variables
measured were level of education and insomnia status (normal sleeper or chronic
insomniac). The researchers discovered that the lower the level of education,
the more likely the person was to have chronic insomnia.

\vspace{0.2cm}

\noindent
\textbf{(a)}[2 points] What are the observational units in this study?

\vspace{1cm} 

\noindent
\textbf{(b)}[3 points] Fill in the blanks with one of the options provided in
parentheses to identify and explain the study design.

This is a(n) (observational study / randomized experiment)
\underline{\hspace{3in}} because (level of education / insomnia status) \underline{\hspace{3in}} (was / was not) \underline{\hspace{3in}} randomly
(assigned / sampled) \underline{\hspace{3in}. 

\vspace{1cm}

\noindent
\textbf{(c)}[3 points] Which types of sampling bias may be present in this
study? Select **all** that are present, or if there is no bias present, select
"No bias." 

\begin{itemize}
\item Selection bias \\
\item Response bias \\
\item Non-response bias \\
\item No bias
\end{itemize}

\newpage

\noindent
\textbf{Q3}[] 
In 2004, the state of North Carolina released to the public a large dataset
containing information on births recorded in this state. This data set has been 
of interest to medical researchers who are studying the relation between habits
and practices of expectant mothers and the birth of their children. This 
analysis will focus on a random sample of 1,000 observations from the published
dataset. 

Our analysis will focus on the relationship between a baby's birth
weight and the mother's smoking status. 

```{r}
births14 
```

\noindent
\textbf{(a)}[2 points] Using the preview of the dataset above, what is the observational 
unit of this study?

\vspace{0.5in}

\noindent
\textbf{(b)}[2 points] Is this an observational study or an experiment? Circle one. 

\begin{center}
\textbf{Observational Study} \hspace{1in} \textbf{Experiment}
\end{center}

\vspace{0.25cm}

\noindent
\textbf{(c)}[3 points] Researchers are interested in the difference in the mean baby birth
weight between mothers who smoke and mothers who do not smoke. Using the table
below, report the observed statistic for this comparison. Indicate in your 
answer what notation should be used for this statistic. 

\vspace{0.25cm}

\textbf{Answer:}

\vspace{0.5in}

\textbf{Notation:}

\vspace{0.25cm}

```{r favstats-smoking}
favstats(weight ~ habit, data = births14)
```

\newpage 

```{r}
births14 %>% 
  ggplot(mapping = aes(x = weight)) + 
  geom_histogram(binwidth = 1.5) + 
  facet_wrap(~habit, scales = "free") + 
  theme_bw() +
  labs(x = "Weight of Baby (lbs)", 
       y = "Number of Babies")
```

\noindent
\textbf{(d)} 



