---
title: "Week 6 Day 2: Comparing Many Means"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Comparing Many Groups


#### Displaying a single categorical and single quantitative variable {-}

The boxplot of movie budgets (in millions) by content rating is plotted using
the code below.  Enter the variable `budget_mil` for `response` and the variable
`content_rating` for explanatory at line 44, highlight and run code lines
42--48. This plot compares the budget for different levels of content rating.

```{r anova-viz, echo = FALSE}
movies %>%  
  filter(content_rating != "Not Rated") %>% 
  ggplot(aes(y = budget_mil, x = content_rating)) + 
  geom_boxplot() + 
  labs(x = "Content Rating", 
       y = "Budget (in Millions)"
       )
```

16. Answer the following questions about the box plots created.

   a. Which content rating has the highest center?

\vspace{0.2in}

   b. Which content rating has the largest spread?

\vspace{0.2in}

   c. Which content rating has the most skewed distribution?

\vspace{0.2in}

   d. Fifty percent of movies in 2016 with a PG-13 content rating fall below
   what value?  What is the name of this value?

\vspace{0.4in}

   e.  What is the value for the third quartile (Q3) for the PG-13 rating?
   Interpret this value in context.

\vspace{.8in}
