---
title: "Midterm 1 -- Group Portion"
author: "Your group's names here!"
format: pdf 
execute: 
  echo: true
  message: false
  warning: false
  error: true
---

```{r setup, include = FALSE}
library(tidyverse)
library(mosaic)
library(moderndive)
```

## Investigating hours worked

For this exam, we'll be working with data from Lab 3 --- the General Social Survey (GSS). The GSS is a high-quality survey which gathers data on American society and opinions, conducted since 1972. This data set is a sample of 500 entries from the GSS, spanning years 1973-2018, including demographic markers and some economic variables.

```{r data-load}
GSS <- read_csv(here::here("Week 5 - Difference in Means & Midterm 1", 
                           "midterm 1",
                           "group", 
                           "data",
                           "gss.csv"
                           )
                )


GSS <- read_csv(here::here("data", 
                           "gss.csv"
                           )
                )
```

## Investigating Education & Spousal Education

You've explored regression before, both in Lab 4 and in your Midterm Project. For this exam, we will investigate if / how much the highest year of school a person's spouse completed (`highest_year_school_completed_spouse`) can predict a person's education (`highest_year_of_school_completed`).

## Summarizing the Variables

**The code below finds summary statistics for the highest year of school each person in this sample completed.**

```{r education-summary}
favstats(~ highest_year_of_school_completed, data = GSS)
```

**1. Using the favstats() function, find summary statistics for the highest year of school each person's spouse completed.**

```{r spouse-summary}
favstats(~ ____, data = GSS)
```

## Visualizing the Relationship

Next, let's visualize the relationship between the highest year of school a person's spouse completed (`highest_year_school_completed_spouse`) and the highest year of school their spouse completed (`highest_year_of_school_completed`). To be clear, `highest_year_school_completed_spouse` is our **explanatory** variable and `highest_year_of_school_completed` is our **response** variable.

**2. I've started the code for you, but you will need to add:**

-   the variables
-   the points
-   nice axis labels

```{r scatterplot}
ggplot(data = GSS, 
       mapping = aes(x = ____, y = ____)) 

# Note: You need to add code to create the points and axis labels!
```

**3. Describe the relationship you see in the plot. Be sure to address the shape, direction and strength of the relationship.**

**4. Are there any outliers present in these data? If so, where?**

**5. If we don't remove the outliers you mentioned, what will happen to the regression line?**

## Linear Regression

Alright, the relationship between `highest_year_of_school_completed` and `highest_year_school_completed_spouse` appears reasonably linear, so let's model this relationship with a line.

**6. Use the lm() function to fit a linear regression between `highest_year_of_school_completed` (response) and `highest_year_of_school_completed` (explanatory).**

```{r lm-fit}



```

**7. Use the get_regression_table() function to obtain the table of coefficient estimates.**

```{r coefficient-table}



```

**8. Using the coefficient table, write out the estimated regression line. Be sure to use the context of the data, not generic x and y statements!**

**9. Interpret the slope coefficient in the context of these data.**

***When you have completed the exam, click the "Render" button to produce your HTML file. Double check to make sure your answers are all there. Then export the HTML file from RStudio Cloud and submit to the Midterm 1 Group Canvas assignment.***
