---
title: "Lab 6: Color Interference"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")

library(tidyverse)
library(mosaic)
library(infer)
```

**Questions associated with the Lab 6 Canvas quiz are noted in boldface.**

### Color Interference

The abstract of the article "Studies of interference in serial verbal reactions"
in the _Journal of Experimental Psychology_ reads:

> In this study pairs of conflicting stimuli, both being inherent aspects of the
same symbols, were presented simultaneously (a name of one color printed in the
ink of another color---a word stimulus and a color stimulus).
> The difference in time for reading the words printed in colors and the same
words printed in black is the measure of interference of color stimuli upon
reading words.
> The interference of conflicting color stimuli upon the time for reading 100
words (each word naming a color unlike the ink-color of its print) caused an
increase of 2.3 seconds or 5.6% over the normal time for reading the same words
printed in black.

The article reports on the results of a study in which seventy college
undergraduates were given forms with 100 names of colors written in black ink,
and the same 100 names of colors written in another color (i.e., the word purple
written in green ink).  The total time (in seconds) for reading the 100 words
printed in black, and the total time (in seconds) for reading the 100 words
printed in different colors were recorded for each subject. The order in which
the forms (black or color) were given was randomized to the subjects. 

We seek to address the following research question: 

__Does printing the name of colors in a different color increase the time it takes to read the words?__

## Identify the scenario

__1. Should these observations be considered paired or independent? Explain your answer.__


__2. Based on your answer to question 1, is the appropriate summary measure to be used to analyze these data the difference in mean times or the mean difference in times?__



## Forming Hypotheses

__3. Write out the null hypothesis in words, in the context of this study.__



__4. Write out the alternative hypothesis in proper notation for this study.__




## Summarize and Visualize the Data

Since the original data from the study are not available, I simulated data to
match the means and standard deviations reported in the article. We will use
these simulated data in the analysis below.

```{r data}
color <- read.csv(here::here("Week 6 - Difference in Means & Paired Data", 
                             "lab", 
                             "data", 
                             "interference.csv")
                  )
```


### Comparison of Letter Color Times

Let's first visualize the distribution of each of the letter color options. 

5. Create histograms of the following: 

- the distribution of reading times for words printed in black 
- the distribution of reading times for words printed in a different color


```{r letter-comparison}
## Code for histogram 1: Reading times printed in black


## Code for histogram 2: Reading times printed in a different color


```


### Visualizing the Differences in Times

If you look at the data, you will notice there are three columns: `Subj`, 
`DiffCol`, and `Black`. There is not, however a column with the differences 
between the different color read time (`DiffCol`) and the black read time
(`Black`). Let's make a new column with the differences!

```{r data-cleaning}
color <- mutate(color,
                Difference = DiffCol - Black
                )
```

Now, let's visualize the distribution of the differences in the reading times. 

```{r difference-histogram}
ggplot(data = color, 
       mapping = aes(x = Difference)) + 
  geom_histogram(binwidth = 5) +
  labs(x = "Differences in Reading Time (Different Color - Black)")
```


### Summarizing the Differences in Times

Our final step is to calculate summary statistics for the differences in 
reading times. Using the `Difference` column created above, write the code to
obtain summary statistics (using the favstats() function) of the differences in
reading time. 

```{r difference-stats}

```


## Check Theoretical Conditions

Yesterday, we explored using simulation-based methods to obtain a p-value 
(permutation) and a confidence interval (bootstrapping). Today, we are going to 
focus on how we could have obtained these using *theoretical* methods instead. 

### The Central Limit Theorem -- A Refresher

In general, the sampling distribution for a sample mean, $\bar{x}$, based on a
sample of size $n$ from a population with a true mean $\mu$ and true standard
deviation $\sigma$ can be modeled using a Normal distribution when certain
conditions are not violated.

Conditions for the sampling distribution of $\bar{x}$ to follow an approximate
Normal distribution:

* **Independence**: The sample’s observations are independent.  For paired data,
that means each *difference* should be independent.

* **Normality**: The data should be approximately normal or the sample size
should be "sufficiently large."

    - $n < 30$: If the sample size $n$ is less than 30 and the distribution of
    the data is approximately normal with no clear outliers in the data, then we
    typically assume the data come from a population with a nearly normal
    distribution to satisfy the condition.

    - $n \geq 30$: If the sample size $n$ is at least 30 and there are **no**
    particularly *extreme* outliers in the data, then we typically assume the
    sampling distribution of $\bar{x}$ is nearly normal, even if the underlying
    distribution of individual observations is not.
    
    - $n \geq 60$: If the sample size $n$ is at least 60 (regardless of the
    presence of skew or outliers), we typically assume the sampling distribution
    of $\bar{x}$ is nearly normal, even if the underlying distribution of
    individual observations is not.
    
 
5. How do you know the independence condition is not violated for these data?


__6. Is the normality condition violated? Would it be inappropriate to use theory-based methods for analysis?  Explain your answer.__



## Use Statistical Methods to Draw Inferences from the Data

### Mathematical Formulas 

To find the standardized statistic for the paired differences we will use the following formula:

$$T = \frac{\bar{x}_d - \mu_d}{SE(\bar{x}_d)},$$
where the standard error of the sample mean difference is:

$$SE(\bar{x}_d)=\frac{s_d}{\sqrt{n}}.$$

__7.  Calculate the standard error of the sample mean difference.__
(Feel free to use the code block below as a calculator)



```{r SE-calc}

```


8. Approximately how many standard errors is the observed mean difference from the null mean difference?



### Using `R` to Obtain a p-value and Confidence Interval

We can use the handy t_test() function from the **infer** package to obtain 
our p-value for testing if the mean difference is different from 0. Luckily, we
can also use this same function to obtain a confidence interval for the mean
difference! 

I've started the code for you, but there are two places you need to fill in 
values. 

- For the blank after *mu*, you need to insert the # we assume the mean
difference equals **in the null hypothesis**. 

- For the blank after *alternative*, you need to insert what direction R should 
use when calculating the p-value. Your three choices are: "greater", "less", or
"two-sided". 

9. Fill in the code with these necessary pieces! 

```{r t-test}
t_test(color, response = Difference, 
       mu = ____, 
       alternative = "____", 
       conf.int = TRUE, 
       conf.level = 0.99
      )
```

10. What does the p-value you obtained mean, in the context of the study? 
*Hint*: it is the probability of what...assuming what?


11. Report the 95% confidence interval for the mean difference. 



12.  Interpret the confidence interval in context of the study.



13.  Do the results of the CI agree with the p-value?  Explain your answer.



14.  The abstract states, that the conflicting color stimuli "caused an increase
of 2.3 seconds or 5.6% over the normal time for reading the same words printed
in black." Is this statement valid? Explain.


